---
title: "Maryland_Final_Figures"
author: "Michelle Bode"
output: html_document
date: "2025-12-15"
note: Must first run scripts "Maryland_ITS_Analysis_Prep_Final.rmd", "Maryland_16S_Analysis_Prep_Final.rmd" and "Maryland_dpcr_analysis_preperation.rmd". 
---

#Set working directory
```{r setup, include=TRUE}
knitr::opts_knit$set(root.dir = "[input pathway to directory with data]")

#Directory for saving figures
f<-"[input pathway to where figure will be saved]"
```

#Activate packages
```{r}
library("phyloseq")
library("ggplot2")
library("patchwork")
library("ggpubr")
library("ALDEx2")
library("phyloseq")
library("dplyr")
library("tidyverse")
library ("BiocManager")
library("devtools")
library("httr")
library("jsonlite") 
library("reticulate")
library("qiime2R")
library("vegan")
library("decontam")
```

#Figure 1A: Bioburden 

#Graph of CHG bioburden dotplot
```{r}
#Order factors correctly
grouped_chg_dpcr$visit_2 <- factor(grouped_chg_dpcr$visit_2, levels = c("Pre-treatment (n=45)","3 Months (n=38)","12 Months (n=31)", "Post-treatment (n=34)"))

y <-"bioburden_chg_dotplot"

#Visualization 

bioburden_dotplot_chg<-ggplot(grouped_chg_dpcr, aes(x=visit_2, y=log_Concentration, group = Patient_id))+
  scale_colour_manual(values=c("Pre-treatment (n=45)"="#6BE0F2","3 Months (n=38)"="#FFF86E","12 Months (n=31)"="#FCC44E","Post-treatment (n=34)"="#69F5AF","Increase" = "#052BEB", "Decrease" ="#D10A0A", "Missing previous visit"="#A8A7A7","No change"="black"), breaks =c("Increase","Decrease","Missing previous visit"), guide = guide_legend(override.aes =list(linewidth =2, size =12)))+
  geom_line(alpha = 0.8, size = 0.6, position=position_dodge(width=0.3), aes(colour=Change_for_segment)) + 
  geom_point(position = position_dodge(width = 0.3), size = 5, aes(colour=visit_2))+
  geom_point(position = position_dodge(width = 0.3), size = 5, colour="black", shape = 21)+
  scale_x_discrete(labels =c(
  "Pre-treatment (n=45)" = "Pre-trt\n(n=45)",  
  "3 Months (n=38)" = " 3 Months\n(n=38)",
  "12 Months (n=31)" = "12 Months\n(n=31)",
  "Post-treatment (n=34)" = "Post-trt\n(n=34)"))+
  theme_classic(base_size = 20)+
  scale_y_continuous(
  limits = c(0.5, 4.5),
  breaks = seq(0, 5, 1),
  expand = expansion(mult = c(0, 0))
)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.position = "bottom",
        legend.title =element_text(size=26))+
labs(title = "", x = "Chlorhexidine treatment", y = expression(bold("log"[10]* "Total 16S Concentration")), colour = "Change from previous visit:")
bioburden_dotplot_chg
```

#Save graph
```{r}
setwd(f)

#Save as image
png(filename = paste0(y, ".png"),
     width = 2400, height = 4500, res = 300)
bioburden_dotplot_chg
dev.off()

#Save plot as pdf
pdf(file = paste0(y, ".pdf"),
    width = 7.73, height = 16.01)
bioburden_dotplot_chg
dev.off()
```


#Graph of water bioburden dotplot
```{r}
#Order factors correctly
grouped_water_dpcr$visit_2 <- factor(grouped_water_dpcr$visit_2, levels = c("Pre-treatment (n=38)","3 Months (n=32)","12 Months (n=28)", "Post-treatment (n=31)"))

y <-"bioburden_water_dotplot"

#Visualization 

bioburden_dotplot_water<-ggplot(grouped_water_dpcr, aes(x=visit_2, y=log_Concentration, group = Patient_id))+
  geom_line(alpha = 0.8, size = 0.6, position=position_dodge(width=0.3), aes(colour=Change_for_segment)) + 
  geom_point(position = position_dodge(width = 0.3), size = 5, aes(colour=visit_2))+
  geom_point(position = position_dodge(width = 0.3), size = 5, colour="black", shape = 21)+
    scale_x_discrete(labels =c(
  "Pre-treatment (n=38)" = "Pre-trt\n(n=38)",  
  "3 Months (n=32)" = " 3 Months\n(n=32)",
  "12 Months (n=28)" = "12 Months\n(n=28)",
  "Post-treatment (n=31)" = "Post-trt\n(n=31)"))+
 scale_colour_manual(values=c("Pre-treatment (n=38)"="#6BE0F2","3 Months (n=32)"="#FFF86E","12 Months (n=28)"="#FCC44E","Post-treatment (n=31)"="#69F5AF","Increase" = "#052BEB", "Decrease" ="#D10A0A", "Missing previous visit"="#A8A7A7","No change"="black"), breaks =c("Increase","Decrease","Missing previous visit"), guide = guide_legend(override.aes =list(linewidth =2, size =12)))+
  theme_classic(base_size = 20)+
  scale_y_continuous(
  limits = c(0.5, 4.5),
  breaks = seq(0, 5, 1),
  expand = expansion(mult = c(0, 0))
)+
  theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.position = "bottom",
        legend.title =element_text(size=26))+
labs(title = "", x = "Soap and water treatment", y = expression(bold("log"[10] * " Total 16S Concentration")), colour = "Change from previous visit:")
bioburden_dotplot_water
```

#save graph
```{r}
setwd(f)

#Save as image
png(filename = paste0(y, ".png"),
    width = 2400, height = 4500, res = 300)
bioburden_dotplot_water
dev.off()

#Save plot as pdf
pdf(file = paste0(y, ".pdf"),
    width = 7.63, height = 16.01)
bioburden_dotplot_water
dev.off()
```

#merge graphs
```{r}
y<-"bioburden_merged_dotplot"
bioburden_merged_dotplot <-bioburden_dotplot_chg +
  bioburden_dotplot_water + 
  plot_layout(guides = "collect") &
theme(legend.position = "bottom",
    legend.margin = margin(t = 15)
  )
```

```{r}
setwd(f)
#Save as image
png(filename = paste0(y, ".png"),
     width = 4400, height = 4500, res = 300)
bioburden_merged_dotplot
dev.off()

#Save plot as pdf
pdf(file = paste0(y, ".pdf"),
    width = 14.66, height = 16.01)
bioburden_merged_dotplot
dev.off()
```


#Figure 1B: Alpha Diversity (Observed)

#CHG alpha diversity 
```{r}
#Isolate treatment group
chg_rare_ps<-ps_rare_genus
meta_data<-as.data.frame(sample_data(ps_rare_genus))
chg_rare_genus<-subset(meta_data, treatgroup == "1")

#Add number of samples to labels
grouped_chg<-chg_rare_genus %>% 
  group_by(visit_2) %>% 
  mutate(n=n()) %>% 
  mutate(visit_2 = paste0(visit_2, " (n=",n,")"))

grouped_chg$sample_id <- rownames(chg_rare_genus)
grouped_chg <- column_to_rownames(grouped_chg, "sample_id")

sample_data(chg_rare_ps) <- grouped_chg

chg_rare_ps<-prune_taxa(taxa_sums(chg_rare_ps)>0,chg_rare_ps)
chg_rare_ps
```

#water alpha diversity 
```{r}
#Isolate treatment group
water_rare_ps<-ps_rare_genus
meta_data<-as.data.frame(sample_data(ps_rare_genus))
water_rare_genus<-subset(meta_data, treatgroup == "0")

#Add number of samples to labels
grouped_water<-water_rare_genus %>% 
  group_by(visit_2) %>% 
  mutate(n=n()) %>% 
  mutate(visit_2 = paste0(visit_2, " (n=",n,")"))

grouped_water$sample_id <- rownames(water_rare_genus)
grouped_water <- column_to_rownames(grouped_water, "sample_id")

sample_data(water_rare_ps) <- grouped_water

water_rare_ps<-prune_taxa(taxa_sums(water_rare_ps)>0,water_rare_ps)

```


#Observed species

#CHG
```{r}
library(ggplot2)
library(ggstatsplot)
library(ggpubr)

#Pick_phyloseq group
ps <-chg_rare_ps

#Order factors correctly
sample_data(ps)$visit_2 <- factor(sample_data(ps)$visit_2, levels = c("Pre-treatment (n=42)","3 Months (n=27)","12 Months (n=25)", "Post-treatment (n=28)"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment (n=42)","3 Months (n=27)"),c("Pre-treatment (n=42)","12 Months (n=25)"),c("Pre-treatment (n=42)","Post-treatment (n=28)"))

#Check if adjusted p-values align with results determined later
rich <- estimate_richness(ps, measures = "Observed")
df <- cbind(as.data.frame(sample_data(ps)), rich)

stats<-compare_means(Observed ~ visit_2,
              data = df,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

#Title of graph 
x <-"Alpha_Diversity_observed_change_overtime_CHG_samples"

#Include n-values
meta<-sample_data(ps)



#Generate plot 
Chg_alpha_diversity_observed<- plot_richness(ps, x="visit_2", title = NULL, measures=c("Observed"), color = "visit_2")+
  geom_jitter(
  aes(fill = visit_2),
  shape = 21,
  color = "black",
  stroke = 0.3,
  width = 0.16,
  size = 3,
  alpha = 1)+
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)+
  scale_x_discrete(labels =c(
  "Pre-treatment (n=42)" = "Pre-trt\n(n=42)",  
  "3 Months (n=27)" = "3 Months\n(n=27)",
  "12 Months (n=25)" = "12 Months\n(n=25)",
  "Post-treatment (n=28)" = "Post-trt\n(n=28)"))+
  theme_classic(base_size = 20)+
  theme_classic(base_size = 20)+
  scale_y_continuous(limits=c(0,270))+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
        legend.position = "none", 
       strip.text = element_blank())+
  scale_color_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     size = 12,
                     label.y = c(190,220,250),
                    comparisons = comparisons)+
  labs(color = "Visit", x = "Chlorhexidine treatment", y = "Observed species richness")

#remove defult points
Chg_alpha_diversity_observed$layers <- Chg_alpha_diversity_observed$layers[-1]

Chg_alpha_diversity_observed
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 6.67, height = 8.1) 
Chg_alpha_diversity_observed
dev.off()

#Save as image
png(filename = paste0(x,".png"), width = 2000, height = 2500, res = 300)
Chg_alpha_diversity_observed 
dev.off()
```

#Average diversity
```{r}
ps<-chg_rare_ps
rich <- estimate_richness(ps, measures = "Observed")
interest_group <- cbind(as.data.frame(sample_data(ps)), rich)


summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean = mean(Observed, na.rm = TRUE),
    sd = sd(Observed, na.rm = TRUE),
    se = sd / sqrt(n),
    ci95_low  = mean - qt(0.975, df = n-1) * se,
    ci95_high = mean + qt(0.975, df = n-1) * se
  )

summary_stats
```


#Water
```{r}

#Pick_phyloseq group
ps <-water_rare_ps

#Order factors correctly
sample_data(ps)$visit_2 <- factor(sample_data(ps)$visit_2, levels = c("Pre-treatment (n=27)","3 Months (n=24)","12 Months (n=22)", "Post-treatment (n=24)"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment (n=27)","3 Months (n=24)"),c("Pre-treatment (n=27)","12 Months (n=22)"),c("Pre-treatment (n=27)","Post-treatment (n=24)"))

#Check if adjusted p-values align with results determined later
rich <- estimate_richness(ps, measures = "Observed")
df <- cbind(as.data.frame(sample_data(ps)), rich)

stats<-compare_means(Observed ~ visit_2,
              data = df,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

#Title of graph 
x <-"Alpha_Diversity_observed_change_overtime_water_samples"

#Include n-values
meta<-sample_data(ps)

#Generate plot 
water_alpha_diversity_observed<- plot_richness(ps, x="visit_2", title = NULL, measures=c("Observed"), color = "visit_2")+
  geom_jitter(
  aes(fill = visit_2),
  shape = 21,
  color = "black",
  stroke = 0.3,
  width = 0.10,
  size = 3,
  alpha = 1)+
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)+
  scale_y_continuous(limits=c(0,270))+
  scale_x_discrete(labels =c(
  "Pre-treatment (n=27)" = "Pre-trt\n(n=27)",  
  "3 Months (n=24)" = "3 Months\n(n=24)",
  "12 Months (n=22)" = "12 Months\n(n=22)",
  "Post-treatment (n=24)" = "Post-trt\n(n=24)"))+
  theme_classic(base_size = 20)+
  theme_classic(base_size = 20)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
        legend.position = "none",
        strip.text = element_blank())+
  scale_color_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     position = "top",
                     size = 12,
                     label.y = c(190,220,250),
                     comparisons = comparisons)+
  labs(color = "Visit", x = "Soap and water treatment", y = "Observed species richness")

#remove defult points
water_alpha_diversity_observed$layers <- water_alpha_diversity_observed$layers[-1]

water_alpha_diversity_observed
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 6.67, height = 8.1) # Adjust dimensions as needed
water_alpha_diversity_observed
dev.off()
#Save as image
png(filename = paste0(x,".png"), width = 2000, height = 2500, res = 300)
water_alpha_diversity_observed
dev.off()
```
#Average diversity
```{r}
ps<-water_rare_ps
rich <- estimate_richness(ps, measures = "Observed")
interest_group <- cbind(as.data.frame(sample_data(ps)), rich)


summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean = mean(Observed, na.rm = TRUE),
    sd = sd(Observed, na.rm = TRUE),
    se = sd / sqrt(n),
    ci95_low  = mean - qt(0.975, df = n-1) * se,
    ci95_high = mean + qt(0.975, df = n-1) * se
  )

summary_stats
```


#Merged
```{r}
#Title of graph 
x <-"Alpha_Diversity_observed_change_overtime_merged_samples"

Alpha_observed_merged<-Chg_alpha_diversity_observed + water_alpha_diversity_observed

Alpha_observed_merged
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 14.50, height =  8.3) # Adjust dimensions as needed
Alpha_observed_merged # Replace with your plot object
dev.off()
#Save as image
png(filename = paste0(x,".png"), width =  4200 , height = 2700, res = 300)
Alpha_observed_merged # Replace with your actual graph object
dev.off()
```

#Figure 1C: Alpha Diversity (Shannon)

#CHG
```{r}
library(ggplot2)
library(ggstatsplot)
library(ggpubr)

#Pick_phyloseq group
ps <-chg_rare_ps

#Order factors correctly
sample_data(ps)$visit_2 <- factor(sample_data(ps)$visit_2, levels = c("Pre-treatment (n=42)","3 Months (n=27)","12 Months (n=25)", "Post-treatment (n=28)"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment (n=42)","3 Months (n=27)"),c("Pre-treatment (n=42)","12 Months (n=25)"),c("Pre-treatment (n=42)","Post-treatment (n=28)"))

#Check if adjusted p-values align with results determined later
rich <- estimate_richness(ps, measures = "Shannon")
df <- cbind(as.data.frame(sample_data(ps)), rich)

stats<-compare_means(Shannon ~ visit_2,
              data = df,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

x <-"Alpha_Diversity_shannon_change_overtime_CHG_samples"

#Include n-values
meta<-sample_data(ps)

#Generate plot 
Chg_alpha_diversity_shannon<- plot_richness(ps, x="visit_2", title = NULL, measures=c("Shannon"), color = "visit_2")+
  geom_jitter(
  aes(fill = visit_2),
  shape = 21,
  color = "black",
  stroke = 0.3,
  width = 0.16,
  size = 3,
  alpha = 1)+
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)+
  scale_y_continuous(limits = c(0,6.5))+
  scale_x_discrete(labels =c(
  "Pre-treatment (n=42)" = "Pre-trt\n(n=42)",  
  "3 Months (n=27)" = "3 Months\n(n=27)",
  "12 Months (n=25)" = "12 Months\n(n=25)",
  "Post-treatment (n=28)" = "Post-trt\n(n=28)"))+
  theme_classic(base_size = 20)+
  theme_classic(base_size = 20)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
        legend.position = "none", 
       strip.text = element_blank())+
  scale_color_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     p.adjust.method = "bonferroni", 
                     size = 12,
                     label.y = c(4.5,5.2,5.9),
                    comparisons = comparisons)+
  labs(color = "Visit", x = "Chlorhexidine treatment", y = "Shannon diversity index")

#remove defult points
Chg_alpha_diversity_shannon$layers <- Chg_alpha_diversity_shannon$layers[-1]

Chg_alpha_diversity_shannon
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 6.67, height = 8.1) 
Chg_alpha_diversity_shannon
dev.off()

#Save as image
png(filename = paste0(x,".png"), width = 2000, height = 2500, res = 300)
Chg_alpha_diversity_shannon
dev.off()
```
#Average diversity
```{r}
ps<-chg_rare_ps
rich <- estimate_richness(ps, measures = "Shannon")
interest_group <- cbind(as.data.frame(sample_data(ps)), rich)


summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean = mean(Shannon, na.rm = TRUE),
    sd = sd(Shannon, na.rm = TRUE),
    se = sd / sqrt(n),
    ci95_low  = mean - qt(0.975, df = n-1) * se,
    ci95_high = mean + qt(0.975, df = n-1) * se
  )

summary_stats
```

#Water
```{r}
library(ggplot2)
library(ggstatsplot)
library(ggpubr)

#Pick_phyloseq group
ps <-water_rare_ps

#Order factors correctly
sample_data(ps)$visit_2 <- factor(sample_data(ps)$visit_2, levels = c("Pre-treatment (n=27)","3 Months (n=24)","12 Months (n=22)", "Post-treatment (n=24)"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment (n=27)","3 Months (n=24)"),c("Pre-treatment (n=27)","12 Months (n=22)"),c("Pre-treatment (n=27)","Post-treatment (n=24)"))

#Check if adjusted p-values align with results determined later
rich <- estimate_richness(ps, measures = "Shannon")
df <- cbind(as.data.frame(sample_data(ps)), rich)

stats<-compare_means(Shannon ~ visit_2,
              data = df,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

x <-"Alpha_Diversity_shannon_change_overtime_water_samples"

#Include n-values
meta<-sample_data(ps)

#Generate plot 
water_alpha_diversity_shannon<- plot_richness(ps, x="visit_2", title = NULL, measures=c("Shannon"), color = "visit_2")+
  geom_jitter(
  aes(fill = visit_2),
  shape = 21,
  color = "black",
  stroke = 0.3,
  width = 0.10,
  size = 3,
  alpha = 1)+
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)+
 scale_y_continuous(limits = c(0,6.5))+
  scale_x_discrete(labels =c(
  "Pre-treatment (n=27)" = "Pre-trt\n(n=27)",  
  "3 Months (n=24)" = "3 Months\n(n=24)",
  "12 Months (n=22)" = "12 Months\n(n=22)",
  "Post-treatment (n=24)" = "Post-trt\n(n=24)"))+
  theme_classic(base_size = 20)+
  theme_classic(base_size = 20)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
        legend.position = "none",
        strip.text = element_blank())+
  scale_color_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     position = "top",
                     size = 12,
                     label.y = c(4.5,5.2,5.9),
                     comparisons = comparisons)+
  labs(color = "Visit", x = "Soap and water treatment", y = "Shannon diversity index")

#remove defult points
water_alpha_diversity_shannon$layers <- water_alpha_diversity_shannon$layers[-1]

water_alpha_diversity_shannon
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 6.67, height = 8.1) # Adjust dimensions as needed
water_alpha_diversity_shannon
dev.off()
#Save as image
png(filename = paste0(x,".png"), width = 2000, height = 2500, res = 300)
water_alpha_diversity_shannon
dev.off()
```
#Average diversity
```{r}
ps<-water_rare_ps
rich <- estimate_richness(ps, measures = "Shannon")
interest_group <- cbind(as.data.frame(sample_data(ps)), rich)


summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean = mean(Shannon, na.rm = TRUE),
    sd = sd(Shannon, na.rm = TRUE),
    se = sd / sqrt(n),
    ci95_low  = mean - qt(0.975, df = n-1) * se,
    ci95_high = mean + qt(0.975, df = n-1) * se
  )

summary_stats
```

#Merged
```{r}
#Title of graph 
x <-"Alpha_Diversity_shannon_change_overtime_merged_samples"

Alpha_shannon_merged<-Chg_alpha_diversity_shannon+water_alpha_diversity_shannon

Alpha_shannon_merged

#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 14.50, height =  8.3)  # Adjust dimensions as needed
Alpha_shannon_merged # Replace with your plot object
dev.off()
#Save as image
png(filename = paste0(x,".png"), width =  4200 , height = 2700, res = 300)
Alpha_shannon_merged # Replace with your actual graph object
dev.off()
```

#Merge alpha diversity plots
```{r}
#Alpha

x <-"Alpha_Diversity_merged_samples"

Alpha_observed_merged_2 <- Alpha_observed_merged + theme(plot.margin = margin(b = 30,
                                                                              t = 30,
                                                                              r = 30,
                                                                              l =30))
Alpha_shannon_merged_2 <- Alpha_shannon_merged + theme(plot.margin = margin(b = 30,
                                            t = 30,
                                            l = 30,
                                            r = 30)) 
alpha_diversity_merged_plot<-Alpha_observed_merged_2/Alpha_shannon_merged_2 
alpha_diversity_merged_plot

#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 14.50, height =  8.3)  # Adjust dimensions as needed
alpha_diversity_merged_plot # Replace with your plot object
dev.off()
#Save as image
png(filename = paste0(x,".png"), width =  4200 , height = 5400, res = 300)
alpha_diversity_merged_plot# Replace with your actual graph object
dev.off()
```


#Merge figure 2 plots
```{r}

x <-"figure_2_merged"
alpha_diversity_merged_plot_2 <- alpha_diversity_merged_plot + theme(plot.margin = margin(b = 30,
                                                                              t = 30,
                                                                              r = 30,
                                                                              l =50))
bioburden_merged_dotplot_2 <- bioburden_merged_dotplot + theme(plot.margin = margin(b = 30,
                                            t = 30,
                                            l = 30,
                                            r = 50)) 
figure_2<- (bioburden_merged_dotplot_2 | alpha_diversity_merged_plot_2) +
  plot_layout(widths = c(1.3,1))


#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 31.333, height =  18) 
figure_2 
dev.off()
#Save as image
png(filename = paste0(x,".png"), width =  9400 , height = 5400, res = 300)
figure_2
dev.off()
```

#Figure 2

#Figure 2A:Bray-Curtis boxplot

#CHG

#Prepare data
```{r}
#transform to relative abundance 
rel_abundance = transform_sample_counts(chg_rare_ps, function(x) x*100/ sum(x))

#Ps object with relative abundance
ps_bds <- rel_abundance 

#calculate bray-distance
bray_dist <- distance(ps_bds, method="bray") %>%
  as.matrix() %>%
  as.data.frame()

#Structure table into rows (FIXED: create bray_long)
bray_long <- rownames_to_column(bray_dist, "Sample_1") %>%
  pivot_longer(
    -Sample_1,
    names_to = "Sample_2",
    values_to = "Bray_Curtis",
    values_transform = list(Bray_Curtis = as.numeric)
  )

#format metadata
meta <- as(sample_data(ps_bds), "data.frame") %>%
  rownames_to_column("Sample_ID")

#FIXED: use bray_long
merged_df <- bray_long %>%
  left_join(meta, by = c("Sample_1" = "Sample_ID"))

#Identify visit 1 sample IDs
v1_list <- merged_df %>%
  filter(vcode_prevfootcomp == "1") %>%
  pull(Sample_1) %>%
  unique()

#Isolate visit 1 comparisons
vist_1 <- merged_df %>%
  filter(Sample_2 %in% v1_list)

#Remove duplicate sample pairs
vist_1 <- vist_1 %>%
  mutate(
    pair_min = pmin(Sample_1, Sample_2),
    pair_max = pmax(Sample_1, Sample_2)
  ) %>%
  distinct(pair_min, pair_max, .keep_all = TRUE)

#Remove same-sample comparisons
vist_1 <- vist_1 %>%
  filter(Sample_1 != Sample_2)

#Save output
chg_v1_BC <- vist_1
chg_v1_BC
```



#Plot figure
```{r}
ps <- chg_v1_BC

#Order factors correctly
ps$visit_2 <- factor(ps$visit_2, levels = c("Pre-treatment (n=42)","3 Months (n=27)","12 Months (n=25)", "Post-treatment (n=28)"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment (n=42)","3 Months (n=27)"),c("Pre-treatment (n=42)","12 Months (n=25)"),c("Pre-treatment (n=42)","Post-treatment (n=28)"))

stats<-compare_means( Bray_Curtis ~ visit_2,
              data =  chg_v1_BC,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

y<-"CHG_BC_Boxplots"

BC_Boxplot_chg<-ggplot(ps, aes(x=visit_2,y=Bray_Curtis, fill= visit_2)) +
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)+
  scale_x_discrete(labels =c(
  "Pre-treatment (n=42)" = "Pre-trt\n(n=42)",  
  "3 Months (n=27)" = "3 Months\n(n=27)",
  "12 Months (n=25)" = "12 Months\n(n=25)",
  "Post-treatment (n=28)" = "Post-trt\n(n=28)"))+
  theme_classic(base_size = 20)+
  theme_classic(base_size = 20)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
        legend.position = "none", 
       strip.text = element_blank())+
  scale_color_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     size = 10,
                     comparisons = comparisons)+
  labs(color = "Visit", x = "Chlorhexidine treatment", y = "Bray-Curtis Dissimilarity")

BC_Boxplot_chg
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y,".pdf"), width = 6.67, height = 8.1) # Adjust dimensions as needed
BC_Boxplot_chg
dev.off()

#Save as image
png(filename = paste0(y,".png"), width = 2000, height = 2500, res = 300)
BC_Boxplot_chg
dev.off()
```

#Average distance
```{r}
interest_group<-chg_v1_BC

summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean_dev = mean(Bray_Curtis, na.rm = TRUE),
    sd_dev = sd(Bray_Curtis, na.rm = TRUE),
    se = sd_dev / sqrt(n),
    ci95_low  = mean_dev - qt(0.975, df = n-1) * se,
    ci95_high = mean_dev + qt(0.975, df = n-1) * se
  )

summary_stats
```

#Water

#Prepare data
```{r}
#transform to relative abundance 
rel_abundance = transform_sample_counts(water_rare_ps, function(x) x*100/ sum(x))

#Ps object with relative abundance
ps_bds <- rel_abundance 

#calculate bray-distance
bray_dist <- distance(ps_bds, method="bray") %>%
  as.matrix() %>%
  as.data.frame()

#Structure table into rows (FIXED: create bray_long)
bray_long <- rownames_to_column(bray_dist, "Sample_1") %>%
  pivot_longer(
    -Sample_1,
    names_to = "Sample_2",
    values_to = "Bray_Curtis",
    values_transform = list(Bray_Curtis = as.numeric)
  )

#format metadata
meta <- as(sample_data(ps_bds), "data.frame") %>%
  rownames_to_column("Sample_ID")

#FIXED: use bray_long
merged_df <- bray_long %>%
  left_join(meta, by = c("Sample_1" = "Sample_ID"))

#Identify visit 1 sample IDs
v1_list <- merged_df %>%
  filter(vcode_prevfootcomp == "1") %>%
  pull(Sample_1) %>%
  unique()

#Isolate visit 1 comparisons
vist_1 <- merged_df %>%
  filter(Sample_2 %in% v1_list)

#Remove duplicate sample pairs
vist_1 <- vist_1 %>%
  mutate(
    pair_min = pmin(Sample_1, Sample_2),
    pair_max = pmax(Sample_1, Sample_2)
  ) %>%
  distinct(pair_min, pair_max, .keep_all = TRUE)

#Remove same-sample comparisons
vist_1 <- vist_1 %>%
  filter(Sample_1 != Sample_2)

#Save output
water_v1_BC <- vist_1
water_v1_BC
```


#Plot figure
```{r}
ps <- water_v1_BC

#Order factors correctly
ps$visit_2 <- factor(ps$visit_2, levels = c("Pre-treatment (n=27)","3 Months (n=24)","12 Months (n=22)", "Post-treatment (n=24)"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment (n=27)","3 Months (n=24)"),c("Pre-treatment (n=27)","12 Months (n=22)"),c("Pre-treatment (n=27)","Post-treatment (n=24)"))

#Check if adjusted p-values align with results determined later
stats<-compare_means( Bray_Curtis ~ visit_2,
              data =  water_v1_BC,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

#Plot
y<-"water_BC_Boxplots"

BC_Boxplot_water<-ggplot(ps, aes(x=visit_2,y=Bray_Curtis, fill= visit_2)) +
  geom_boxplot(aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)+
labs(x ="Soap and water treatment", y = "Bray-Curtis Dissimilarity", title = "")+
  theme_classic()+
coord_cartesian(clip = "off") +
  scale_x_discrete(labels =c(
  "Pre-treatment (n=27)" = "Pre-trt\n(n=27)",  
  "3 Months (n=24)" = "3 Months\n(n=24)",
  "12 Months (n=22)" = "12 Months\n(n=22)",
  "Post-treatment (n=24)" = "Post-trt\n(n=24)"))+
  theme_classic(base_size = 20)+
  theme_classic(base_size = 20)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
        legend.position = "none",
        strip.text = element_blank())+
  scale_color_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     size = 10,
                     comparisons = comparisons)

BC_Boxplot_water
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y,".pdf"), width = 6.67, height = 8.1) # Adjust dimensions as needed
BC_Boxplot_water
dev.off()

#Save as image
png(filename = paste0(y,".png"), width = 2000, height = 2500, res = 300)
BC_Boxplot_water 
dev.off()

```
#Average distance
```{r}
interest_group<-water_v1_BC

summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean_dev = mean(Bray_Curtis, na.rm = TRUE),
    sd_dev = sd(Bray_Curtis, na.rm = TRUE),
    se = sd_dev / sqrt(n),
    ci95_low  = mean_dev - qt(0.975, df = n-1) * se,
    ci95_high = mean_dev + qt(0.975, df = n-1) * se
  )

summary_stats
```

#Combine Graphs
```{r}
x<-"combined_betadiversity_boxplots"

# p1 has legend removed
p1 <- BC_Boxplot_chg + 
  theme(legend.position = "none") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.margin = margin(b = 30))

# p2 keeps legend
p2 <- BC_Boxplot_water + 
   theme(legend.position = "none",legend.margin = margin(t = 15), plot.margin = margin(t = 30)
  )+
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

combined_betadiversity_boxplots <- p1 / p2 

combined_betadiversity_boxplots

```

```{r}
#Save plot as pdf

setwd(f)
pdf(file = paste0(x,".pdf"), width = 10.333, height =  7.7) 
combined_betadiversity_boxplots
dev.off()

#Save as image
png(filename = paste0(x,".png"), width =  3000 , height = 5700, res = 300)
combined_betadiversity_boxplots
dev.off()

```


#Figure 3B:Change during treatment

#CHG
#Prepare data
```{r}
#transform to relative abundance 
rel_abundance = transform_sample_counts(chg_rare_ps, function(x) x*100/ sum(x))

#Ps object with relative abundance
ps_bds<-rel_abundance 
ps_bds

#set testing parameter
x="treatment_active"

#Import tree
phy_tree(ps_bds)<-qza_to_phyloseq(tree="/Users/michellebode/maryland/16S/processed_data/tree/rooted_tree.qza")
ps_bds

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot

print("permanova")
permanova_result <- vegan::adonis2(bray_dist ~ treatment_active, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)

bray_active_treat_chg<-ps_bds
```

#Graph
```{r}
y<-"Bray_Curtis_Active_vs_No_Treatetment_chg"

# Run NMDS 
ordination_result <- ordinate(bray_active_treat_chg, method='NMDS', distance='bray', k=4)

#Check stress levels if >0.2 increase k. ideally <0.1
ordination_result$stress

# Plot Axis 1 and Axis 2
bray_active_treat_chg_plot<-plot_ordination(bray_active_treat_chg, ordination_result, axes=c(1,2), color=x)+
geom_point(size = 6) +
  geom_point(
    size = 6,
    shape = 21,        
    color = "black",   
    stroke = 0.3) +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#F54E4E","#8953CF"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
))) +
  theme_classic()+
  theme_bw()+
     theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
       legend.position = "bottom") +
  scale_color_manual(
  values = c(
    "Active treatment (3 months + 12 months)" = "#F54E4E",
    "No active treatment (pre-trt + post-trt" = "#8953CF"
  ),
  labels = c(
    "Active treatment (3 months + 12 months)" = "Active trt (3 & 12 months)",
    "No active treatment (pre-trt + post-trt" = "No active trt (pre & post-trt)"
  )
)+
  labs( title = "", color = "" )

bray_active_treat_chg_plot
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 8) # Adjust dimensions as needed
bray_active_treat_chg_plot
dev.off()

#Save plot as pdf
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 2000, res = 300)
bray_active_treat_chg_plot
dev.off()
```


#Water

#Prepare data
```{r}
#Ps object with exact counts
ps_bds<-water_rare_ps

#created grouped treatement
meta<-sample_data(ps_bds)
meta$current_treat <- ifelse(meta$vcode_prevfootcomp %in% c("1","7"),"No Active treatment","Active Treatment")
sample_data(ps_bds) <- meta

#set testing parameter
x="treatment_active"

#Import tree
phy_tree(ps_bds)<-qza_to_phyloseq(tree="/Users/michellebode/maryland/16S/processed_data/tree/rooted_tree.qza")
ps_bds

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot

print("permanova")
permanova_result <- vegan::adonis2(bray_dist ~ treatment_active, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)

bray_active_treat_water<-ps_bds
```


#Graph
```{r}
y<-"Bray_Curtis_Active_vs_No_Treatetment_water"

# Run NMDS 
ordination_result <- ordinate(bray_active_treat_water, method='NMDS', distance='bray', k=4)

#Check stress levels if >0.2 increase k. ideally <0.1
ordination_result$stress

# Plot Axis 1 and Axis 2
bray_active_treat_water_plot<-plot_ordination(bray_active_treat_water, ordination_result, axes=c(1,2), color=x)+
geom_point(size = 6) +
  geom_point(
    size = 6,
    shape = 21,        
    color = "black",   
    stroke = 0.3) +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#F54E4E","#8953CF"), 
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
))) +
  theme_classic()+
  theme_bw()+
     theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
       legend.position = "bottom") +
scale_color_manual(
  values = c(
    "Active treatment (3 months + 12 months)" = "#F54E4E",
    "No active treatment (pre-trt + post-trt" = "#8953CF"
  ),
  labels = c(
    "Active treatment (3 months + 12 months)" = "Active trt (3 & 12 months)",
    "No active treatment (pre-trt + post-trt" = "No active trt (pre & post-trt)"
  )
)+
  labs( title = "", color = "" )

bray_active_treat_water_plot
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 8) # Adjust dimensions as needed
bray_active_treat_water_plot
dev.off()

#Save plot as pdf
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 2000, res = 300)
bray_active_treat_water_plot
dev.off()
```

#Combine Graphs
```{r}

x<-"combined_treatment_active"

p1 <- bray_active_treat_chg_plot + 
  theme(legend.position = "none")+
  labs(title = "Chlorhexidine treatment") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
    plot.margin = margin(b = 30))

p2 <- bray_active_treat_water_plot + 
  theme( legend.position = "bottom", legend.margin = margin(t = 15), plot.margin = margin(t = 30)) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))+
     labs(title = "Soap and water treatment")

combined_active_treatment <- p1 / p2 

```

```{r}
#Save plot as pdf

setwd(f)
pdf(file = paste0(x,".pdf"), width = 10.333, height =  7.5) 
combined_active_treatment
dev.off()

#Save as image
png(filename = paste0(x,".png"), width =  3000 , height = 5400, res = 300)
combined_active_treatment
dev.off()
```


#Figure 3C: Shift from visit 1 to visit 6

#isolate treatment groups
```{r}
v1v6<-subset_samples(ps_rare_genus, vcode_prevfootcomp %in% c("1","6"))
ps_rare_water<-subset_samples(v1v6, treatgroup == "0")
ps_rare_chg<-subset_samples(v1v6, treatgroup == "1")

# Keep only Patient_id that have both visit 1 and 6
meta_c<-as.data.frame(sample_data(ps_rare_chg))
valid_patients_chg <- meta_c %>%
  group_by(Patient_id) %>%
  filter(all(c("1", "6") %in% vcode_prevfootcomp)) %>%   # visit is probably character, so use "1","6"
  pull(Patient_id) %>%                      # <- extract as vector
  unique()

ps_rare_chg <- prune_samples(sample_names(ps_rare_chg)[sample_data(ps_rare_chg)$Patient_id %in% valid_patients_chg],ps_rare_chg)

meta_w<-as.data.frame(sample_data(ps_rare_water))
valid_patients_water <- meta_w %>%
  group_by(Patient_id) %>%
  filter(all(c("1", "6") %in% vcode_prevfootcomp)) %>%   # visit is probably character, so use "1","6"
  pull(Patient_id) %>%                      # <- extract as vector
  unique()

ps_rare_water <- prune_samples(sample_names(ps_rare_water)[sample_data(ps_rare_water)$Patient_id %in% valid_patients_water],ps_rare_water)

ps_rare_chg
ps_rare_water

#determine number of patients
length(unique(sample_data(ps_rare_chg)$Patient_id))
length(unique(sample_data(ps_rare_water)$Patient_id))
```



#CHG

#Prepare data
```{r}
#transform to relative abundance 
rel_abundance = transform_sample_counts(ps_rare_chg, function(x) x*100/ sum(x))

#Ps object with relative abundance
ps_bds<-rel_abundance 

#Ps object with exact counts
#ps_bds<-unique_genus

#set testing parameter
x="vcode_prevfootcomp"

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

print("permanova")
#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot
bray_dist <- distance(ps_bds, method = "bray")
permanova_result <- vegan::adonis2(bray_dist ~ vcode_prevfootcomp, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)

```

#Graphing differences
```{r}
y <- "Bray_Curtis_NMDS_Chg_1v6"

#order
meta <-as.data.frame(sample_data(ps_bds))
meta$visit <- factor(meta$visit, levels = c("1", "6"), ordered = TRUE)
sample_data(ps_bds) <- meta

# Run NMDS 
ordination_result <- ordinate(ps_bds, method = 'NMDS', distance = 'bray', k = 4)

# Check stress
ordination_result$stress

# Plot Axis 1 and Axis 2
Bray_Curtis_NMDS_Chg_1v6<- plot_ordination(ps_bds, ordination_result, axes = c(1,2), color = "visit_2") +
  geom_point(size =6, 
             alpha =1)+
  geom_point(
    size = 6,
    alpha = 1,
    shape = 21,
    stroke = 0.3,
    color ="black") +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#6BE0F2","#FCC44E"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
  )))+
  geom_path(
    aes(group = Patient_id),
    data = function(df) df[order(df$Patient_id, df$visit_2), ],
    arrow = arrow(type = "closed", length = unit(0.2, "inches")),
    linewidth = 0.4,
    color = "black",
    alpha = 0.9
  ) +
 theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
       legend.position = "none",
        legend.text =element_text(size=26,face ="italic"),
       legend.title = element_text(size=26))+
  labs(title = "", color = "")+
  scale_color_manual(values = c("Pre-treatment"="#6BE0F2", "12 Months"="#FCC44E"))


Bray_Curtis_NMDS_Chg_1v6
```

#print graphs
```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 6) # Adjust dimensions as needed
Bray_Curtis_NMDS_Chg_1v6 # Replace with your plot object
dev.off()

#Save plot as pdf
setwd(f)
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 1500, res = 300)
Bray_Curtis_NMDS_Chg_1v6  # Replace with your actual graph object
dev.off()
```


#Water

#Prepare data
```{r}
#transform to relative abundance 
rel_abundance = transform_sample_counts(ps_rare_water, function(x) x*100/ sum(x))

#Ps object with relative abundance
ps_bds<-rel_abundance 

#Ps object with exact counts
#ps_bds<-unique_genus

#set testing parameter
x="vcode_prevfootcomp"

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

print("permanova")
#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot
bray_dist <- distance(ps_bds, method = "bray")
permanova_result <- vegan::adonis2(bray_dist ~ vcode_prevfootcomp, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)
```

#Graphing differences
```{r}
y <- "Bray_Curtis_NMDS_water_1v6"

#order
meta <-as.data.frame(sample_data(ps_bds))
meta$visit <- factor(meta$visit, levels = c("1", "6"), ordered = TRUE)
sample_data(ps_bds) <- meta

# Run NMDS 
ordination_result <- ordinate(ps_bds, method = 'NMDS', distance = 'bray', k = 4)

# Check stress
ordination_result$stress

# Plot Axis 1 and Axis 2
Bray_Curtis_NMDS_water_1v6<- plot_ordination(ps_bds, ordination_result, axes = c(1,2), color = "visit_2") +
  geom_point(size =6, 
             alpha =1)+
  geom_point(
    size = 6,
    alpha = 1,
    shape = 21,
    stroke = 0.3,
    color ="black") +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#6BE0F2","#FCC44E"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
  )))+
  geom_path(
    aes(group = Patient_id),
    data = function(df) df[order(df$Patient_id, df$visit_2), ],
    arrow = arrow(type = "closed", length = unit(0.2, "inches")),
    linewidth = 0.4,
    color = "black",
    alpha = 0.9
  ) +
 theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
       legend.position = "none",
       legend.title = element_text(size=26),
        legend.text =element_text(size=26,face ="italic"))+
  labs(title = "", color = "")+
  scale_color_manual(values = c("Pre-treatment"="#6BE0F2", "12 Months"="#FCC44E"))


Bray_Curtis_NMDS_water_1v6
```

#print graphs
```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 6) 
Bray_Curtis_NMDS_water_1v6
dev.off()

#Save plot as pdf
setwd(f)
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 1500, res = 300)
Bray_Curtis_NMDS_water_1v6
dev.off()
```

#Combine Graphs
```{r}
x<-"combined_v1_v6"

# p1 has legend removed
p1 <- Bray_Curtis_NMDS_Chg_1v6 + 
  theme(legend.position = "none") + 
  labs(title = "Chlorhexidine treatment") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.margin = margin(b = 30))

# p2 keeps legend
p2 <- Bray_Curtis_NMDS_water_1v6 + 
   theme(legend.position = "bottom",legend.margin = margin(t = 15), plot.margin = margin(t = 30)
  )+
  labs(title = "Soap and water treatment") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

combined_v1_v6 <- p1 / p2 

combined_v1_v6

```

```{r}
#Save plot as pdf

setwd(f)
pdf(file = paste0(x,".pdf"), width = 10.333, height =  7.5) 
combined_v1_v6
dev.off()

#Save as image
png(filename = paste0(x,".png"), width =  3000 , height = 5400, res = 300)
combined_v1_v6
dev.off()
```


#Merge figure 3 plots
```{r}
x <-"figure_3_merged"


combined_betadiversity_boxplots_2 <- combined_betadiversity_boxplots + theme(plot.margin = margin(b = 0,
                                                                              t = 50,
                                                                              r = 30,
                                                                              l =50))


combined_active_treatment_2 <- combined_active_treatment + theme(plot.margin = margin(b = 30,
                                            t = 30,
                                            l = 50,
                                            r = 50)) 

combined_v1_v6_2 <- combined_v1_v6 + theme(plot.margin = margin(b = 30,
                                            t = 30,
                                            l = 50,
                                            r = 50)) 

figure_3 <- (
  combined_betadiversity_boxplots_2 | 
  combined_active_treatment |
  combined_v1_v6_2 
) + 
  plot_layout(widths = c(1,1.18,1))


#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 31.333, height =  18) 
figure_3 
dev.off()
#Save as image
png(filename = paste0(x,".png"), width =  9400 , height = 5400, res = 300)
figure_3 
dev.off()
```





#Figure 4

#Figure 4A: Relative abundance graphs  

#Prepare data
```{r}
#Filter out low features (joined)
ps_rare_water<-subset_samples(ps_nr_genus, treatgroup == "0")
ps_rare_chg<-subset_samples(ps_nr_genus, treatgroup == "1")

cleaned_otu_chg_1<-otu_table(ps_rare_chg)
cleaned_otu_water_1<-otu_table(ps_rare_water)

#remove features with few counts in Chg
cleaned_otu_chg<- cleaned_otu_chg_1[rowSums(cleaned_otu_chg_1) > 20,]

print("Chg features lost")
lost_h=nrow(otu_table(cleaned_otu_chg_1))-nrow(cleaned_otu_chg)
lost_h
lost_h/nrow(cleaned_otu_chg_1)*100

#remove features with few counts in water
cleaned_otu_water<-cleaned_otu_water_1[rowSums(cleaned_otu_water_1) > 20,]

print("water features lost")
lost_h=(nrow(cleaned_otu_water_1)-nrow(cleaned_otu_water))
lost_h
lost_h/nrow(cleaned_otu_water_1)*100

#Check stats 
print("chg samples count distribution")
summary(rowSums(cleaned_otu_chg))
print("water samples count distribution")
summary(rowSums(cleaned_otu_water))

#update phyloseq object 
clean_chg_ps<-ps_rare_chg
clean_water_ps<-ps_rare_water
otu_table(clean_chg_ps)<-cleaned_otu_chg
otu_table(clean_water_ps)<-cleaned_otu_water

#Transform counts
# Transform counts to relative abundances (chg)
ps_rel_abundance_c = transform_sample_counts(clean_chg_ps, function(x) x*100 / sum(x))
# Transform counts to relative abundances (water)
ps_rel_abundance_w = transform_sample_counts(clean_water_ps, function(x) x*100 / sum(x))

#Merge ps into df
ps_tab_genus_df_c = psmelt(ps_rel_abundance_c)
ps_tab_genus_df_w = psmelt(ps_rel_abundance_w)


#Identify top abundant

# Summarize total abundance for each genus
top_genera_c <- ps_tab_genus_df_c %>%
  group_by(Genus) %>%
  summarize(TotalAbundance = sum(Abundance, na.rm = TRUE)) %>%
  arrange(desc(TotalAbundance)) %>%
  slice_head(n = 15)  # Select the top 15 genera

top_genus_names_c<-top_genera_c$Genus

#order highest to lowest
top_genera_c <- top_genera_c %>%
  mutate(Genus = factor(Genus, levels = Genus))

#plot overall most abundant
healthy_plot<-ggplot(top_genera_c, aes(x=Genus, y=TotalAbundance))+
  geom_bar(stat="identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(title="Top 15 abundant genus in chg samples", x = "Genus", y = "Added relative abundances")
healthy_plot

# Summarize total abundance for each genus
top_genera_w <- ps_tab_genus_df_w %>%
  group_by(Genus) %>%
  summarize(TotalAbundance = sum(Abundance, na.rm = TRUE)) %>%
  arrange(desc(TotalAbundance)) %>%
  slice_head(n = 15)  # Select the top 10 genera

top_genus_names_w<-top_genera_w$Genus

#order highest to lowest
top_genera_w <- top_genera_w %>%
  mutate(Genus = factor(Genus, levels = Genus))

#plot overall most abundant
wound_plot<-ggplot(top_genera_w, aes(x=Genus, y=TotalAbundance))+
  geom_bar(stat="identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(title="Top 15 abundant genus in water samples", x = "Genus", y = "Added relative abundances")
wound_plot
```



#Prepare for figures
```{r}
#Water

# Convert Genus to character (if it's a factor)
ps_tab_genus_df_w$Genus <- as.character(ps_tab_genus_df_w$Genus)


# Rename genera not in the list to "Other"
ps_tab_genus_df_w$Genus[!ps_tab_genus_df_w$Genus %in% top_genus_names_w] <- "A_Other"

# Check the unique genera and their count
unique(ps_tab_genus_df_w$Genus)
length(unique(ps_tab_genus_df_w$Genus))

#Chg

# Convert Genus to character (if it's a factor)
ps_tab_genus_df_c$Genus <- as.character(ps_tab_genus_df_c$Genus)

# Rename genera not in the list to "Other"
ps_tab_genus_df_c$Genus[!ps_tab_genus_df_c$Genus %in% top_genus_names_c] <- "A_Other"

#rename shorthand of genera
ps_tab_genus_df_c$Genus <- forcats::fct_recode(
  ps_tab_genus_df_c$Genus,
  "Methylobacterium" = "Methylobacterium-Methylorubrum"
)

# Check the unique genera and their count
unique(ps_tab_genus_df_c$Genus)
length(unique(ps_tab_genus_df_c$Genus))


#Relative Abundance plots

library(dplyr)

# Merge sequentially with base R
colour_df <- vctrs::vec_c(ps_tab_genus_df_w,ps_tab_genus_df_c)

# Summarize by Genus
colour_complete_df <- colour_df %>% group_by(Genus)
length(unique(colour_complete_df$Genus))


#Colour

# Load necessary libraries
library(RColorBrewer)

# Ensure orderus is a factor
colour_complete_df$Genus <- as.factor(colour_complete_df$Genus)


# Define a custom color palette
custom_colors <- c( "#D9D9D9",
 #pinks
  "#D81B60", "#EC407A","#F06292", "#F48FB1",
  #"#F8BBD0",
 
   # Oranges and Yellows
  "#FF6E40","#FF9800","#FFC107","#FFEB3B",
  #"#FFF59D",

  
  # Greens
  #"#1B5E20", 
  "#2E7D32","#4CAF50", "#AED581", "#C8E6C9", 
  

  # Blues and Teals
  "#01579B","#1976D2", "#2196F3", "#64B5F6",
  #"#BBDEFB", 
  
 
  # Purples and Violets
  "#673AB7","#B39DDB","#E6DBF2")

  
  
# Create a named vector for the custom colors
names(custom_colors) <- levels(colour_complete_df$Genus)

# Get the number of distinct levels (orderus)
num_levels <- length(levels(colour_complete_df$Genus))

# Use only as many colors as you have levels
rel_ps_genus_colors <- custom_colors[1:num_levels]

rel_ps_genus_colors
```


#Chg
```{r}
ordered_samples <- ps_tab_genus_df_c%>%
  filter(Genus == "Staphylococcus") %>%
  arrange(desc(Abundance)) %>%
  pull(Sample)

ps_tab_genus_df_c$Sample <- factor(ps_tab_genus_df_c$Sample,
                                     levels = c(ordered_samples,
                                                setdiff(unique(ps_tab_genus_df_c$Sample),
                                                        ordered_samples)))

y<-"Chg_Samples_Relative_Abundance_Plot_Top_Genus_all_outcome"

#Create Plot all samples
Chg_genus_plot <-ggplot(ps_tab_genus_df_c, aes(x=Sample, y=Abundance, fill = Genus)) +
  geom_bar(stat="identity", aes(fill=Genus))+
  labs(x="Chlorhexidine treatment", y="% Relative Abundance", title="", fill ="Top 15 Genera")+
  facet_grid(~visit_2, space="free", scales="free_x", labeller = labeller(
             visit_2 = c(
  "Pre-treatment" = "Pre-trt\n(n=42)",  
  "3 Months" = "3 Months\n(n=27)",
  "12 Months" = "12 Months\n(n=25)",
  "Post-treatment" = "Post-trt\n(n=28)")))+
theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26,face = "bold"),
        axis.title.x.bottom = element_text(size=26,face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.position = "right",
        legend.text =element_text(size=20,face ="italic"),
       legend.title =element_text(size=24))+
  theme(strip.background=element_blank(),
        axis.text.x.bottom = element_blank())+
  scale_fill_manual(values=rel_ps_genus_colors)
Chg_genus_plot

#Save plot as pdf
setwd(f)
pdf(file = paste0(y,".pdf"), width = 15, height = 8) # Adjust dimensions as needed
Chg_genus_plot # Replace with your plot object
dev.off()

#Save as image
setwd(f)
png(filename = paste0(y,".png"), width = 4000, height = 1750, res = 300)
Chg_genus_plot # Replace with your actual graph object
dev.off()
```

#Water
```{r}
ordered_samples <- ps_tab_genus_df_w %>%
  filter(Genus == "Staphylococcus") %>%
  arrange(desc(Abundance)) %>%
  pull(Sample)

ps_tab_genus_df_w$Sample <- factor(ps_tab_genus_df_w$Sample,
                                     levels = c(ordered_samples,
                                                setdiff(unique(ps_tab_genus_df_w$Sample),
                                                        ordered_samples)))

y<-"Water_Samples_Relative_Abundance_Plot_Top_Genus_all_outcome"

#Create Plot all samples
Water_genus_plot <-ggplot(ps_tab_genus_df_w, aes(x=Sample, y=Abundance, fill = Genus)) +
  geom_bar(stat="identity", aes(fill=Genus))+
  labs(x="Soap and water treatment", y="% Relative Abundance", title="", fill ="Top 15 Genera")+
  facet_grid(~visit_2, space="free", scales="free_x", labeller = labeller(
             visit_2 = c(
  "Pre-treatment" = "Pre-trt\n(n=27)",  
  "3 Months" = "3 Months\n(n=24)",
  "12 Months" = "12 Months\n(n=22)",
  "Post-treatment" = "Post-trt\n(n=24)")))+
theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26,face = "bold"),
        axis.title.x.bottom = element_text(size=26, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.position = "right",
        legend.text =element_text(size=20,face ="italic"),
       legend.title =element_text(size=24))+
  theme(strip.background=element_blank(),
        axis.text.x.bottom = element_blank())+
  scale_fill_manual(values=rel_ps_genus_colors)
Water_genus_plot

#Save plot as pdf
setwd(f)
pdf(file = paste0(y,".pdf"), width = 15, height = 8) # Adjust dimensions as needed
Water_genus_plot # Replace with your plot object
dev.off()

#Save as image
setwd(f)
png(filename = paste0(y,".png"), width = 4000, height = 1750, res = 300)
Water_genus_plot # Replace with your actual graph object
dev.off()
```

#merged relative abundance plots
```{r}
#Combine Graphs

x<-"combined_relative_abundance"

# p1 has legend removed
p1 <-  Chg_genus_plot +  
  theme(plot.margin = margin(b =1 ,t=1))

# p2 keeps legend
p2 <- Water_genus_plot + 
   theme(plot.margin = margin(t =1, b=1),  

    # Align legend by the boxes, not the text
    legend.text.align = 0,      # left-align text AFTER the box
    legend.justification = "left",
    legend.box.just = "left"
  )

combined_relative_abundance <- p1 / p2 
combined_relative_abundance

#Save plot as pdf

setwd(f)
pdf(file = paste0(x,".pdf"), width = 15, height = 8) 
combined_relative_abundance
dev.off()

png(filename = paste0(x,".png"), width = 4000, height = 3500, res = 300)
combined_relative_abundance
dev.off()
```


#Figure 4B: Staphyloccus relative abundance

#Prepare data
```{r}
#Isolate Staphylococcus
ps_staph_water<-subset(ps_tab_genus_df_w, Genus=="Staphylococcus") 
ps_staph_chg<-subset(ps_tab_genus_df_c, Genus=="Staphylococcus") 

#Record change in concentration from prior visit
df_staph_water <- ps_staph_water %>%
  group_by(Patient_id) %>%
  arrange(Patient_id, vcode_prevfootcomp) %>%
  mutate(
    prev_code  = lag(vcode_prevfootcomp),
    prev_value = lag(Abundance),
    has_1 = any(vcode_prevfootcomp == 1),
    is_consecutive = case_when(
      vcode_prevfootcomp == 1 ~ NA,
      vcode_prevfootcomp == 3 & has_1 ~ TRUE,
      vcode_prevfootcomp == 3 & !has_1 ~ FALSE,
      vcode_prevfootcomp == 6 & prev_code == 3 ~ TRUE,
      vcode_prevfootcomp == 6 & prev_code == 1 ~ FALSE,
      vcode_prevfootcomp == 6 & is.na(prev_code) ~ FALSE,
      vcode_prevfootcomp == 7 & prev_code == 6 ~ TRUE,
      vcode_prevfootcomp == 7 & prev_code %in% c(1,3) ~ FALSE,
      vcode_prevfootcomp == 7 & is.na(prev_code) ~ FALSE,
      TRUE ~ NA
    ),
    Patient_change = case_when(
      is_consecutive == TRUE & Abundance > prev_value ~ "Increase",
      is_consecutive == TRUE & Abundance < prev_value ~ "Decrease",
      is_consecutive == TRUE & Abundance == prev_value ~ "No change",
      is_consecutive == FALSE ~ "Missing previous visit",
      TRUE ~ NA_character_        
    ),
    Change_for_segment = lead(Patient_change),
    change_amount = Abundance - prev_value
  ) %>%
  ungroup()


df_staph_chg <- ps_staph_chg %>%
  group_by(Patient_id) %>%
  arrange(Patient_id, vcode_prevfootcomp) %>%
  mutate(
    prev_code  = lag(vcode_prevfootcomp),
    prev_value = lag(Abundance),
    has_1 = any(vcode_prevfootcomp == 1),
    is_consecutive = case_when(
      vcode_prevfootcomp == 1 ~ NA,
      vcode_prevfootcomp == 3 & has_1 ~ TRUE,
      vcode_prevfootcomp == 3 & !has_1 ~ FALSE,
      vcode_prevfootcomp == 6 & prev_code == 3 ~ TRUE,
      vcode_prevfootcomp == 6 & prev_code == 1 ~ FALSE,
      vcode_prevfootcomp == 6 & is.na(prev_code) ~ FALSE,
      vcode_prevfootcomp == 7 & prev_code == 6 ~ TRUE,
      vcode_prevfootcomp == 7 & prev_code %in% c(1,3) ~ FALSE,
      vcode_prevfootcomp == 7 & is.na(prev_code) ~ FALSE,
      TRUE ~ NA
    ),
    Patient_change = case_when(
      is_consecutive == TRUE & Abundance > prev_value ~ "Increase",
      is_consecutive == TRUE & Abundance < prev_value ~ "Decrease",
      is_consecutive == TRUE & Abundance == prev_value ~ "No change",
      is_consecutive == FALSE ~ "Missing previous visit",
      TRUE ~ NA_character_        
    ),
    Change_for_segment = lead(Patient_change),
    change_amount = Abundance - prev_value
  ) %>%
  ungroup()
```


#Graph of CHG staph
```{r}
#Order factors correctly
df_staph_chg$visit_2 <- factor(df_staph_chg$visit_2, levels = c("Pre-treatment","3 Months","12 Months","Post-treatment"))

y <-"staph_chg_dotplot"

#Visualization 

library(ggplot2)

staph_dotplot_chg<-ggplot(df_staph_chg, aes(x=visit_2, y=Abundance, group = Patient_id))+
  scale_fill_manual(values=c("Pre-treatment"="#6BE0F2","3 Months"="#FFF86E","12 Months"="#FCC44E","Post-treatment"="#69F5AF"))+
  geom_line(alpha = 0.8, size = 0.6, position=position_dodge(width=0.3), aes(colour=Change_for_segment)) + 
  scale_colour_manual(values=c("Pre-treatment"="#6BE0F2","3 Months"="#FFF86E","12 Months"="#FCC44E","Post-treatment"="#69F5AF","Increase" = "#052BEB", "Decrease" ="#D10A0A", "Missing previous visit"="#A8A7A7","No change"="black"), breaks =c("Increase","Decrease","Missing previous visit"), guide = guide_legend(override.aes =list(linewidth =2, size = 12)))+
  geom_point(position = position_dodge(width = 0.3), size = 5, aes(colour=visit_2))+
  scale_x_discrete(labels =c(
  "Pre-treatment" = "Pre-trt\n(n=42)",  
  "3 Months" = "3 Months\n(n=27)",
  "12 Months" = "12 Months\n(n=25)",
  "Post-treatment" = "Post-trt\n(n=28)"))+
  geom_point(position = position_dodge(width = 0.3), size = 5, colour="black", shape = 21)+
  theme_classic(base_size = 26)+
  scale_y_continuous(limits =c(0,100))+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.position = "bottom",
        legend.title =element_text(size=26))+
labs(title = "", x = "Chlorhexidine treatment", y = "Relative abundance of Staphylococcus %", colour = "Change from previous visit:")
staph_dotplot_chg

setwd(f)

#Save as image
png(filename = paste0(y, ".png"),
     width = 2400, height = 4500, res = 300)
staph_dotplot_chg
dev.off()

#Save plot as pdf
pdf(file = paste0(y, ".pdf"),
    width = 7.73, height = 16.01)
staph_dotplot_chg
dev.off()
```


```{r}
#Order factors correctly
df_staph_water$visit_2 <- factor(df_staph_water$visit_2, levels = c("Pre-treatment","3 Months","12 Months", "Post-treatment"))

y <-"staph_water_dotplot"

#Visualization 

library(ggplot2)

staph_dotplot_water<-ggplot(df_staph_water, aes(x=visit_2, y=Abundance, group = Patient_id))+
   scale_fill_manual(values=c("Pre-treatment"="#6BE0F2","3 Months"="#FFF86E","12 Months"="#FCC44E","Post-treatment"="#69F5AF"))+
  geom_line(alpha = 0.8, size = 0.6, position=position_dodge(width=0.3), aes(colour=Change_for_segment)) + 
  scale_colour_manual(values=c("Pre-treatment"="#6BE0F2","3 Months"="#FFF86E","12 Months"="#FCC44E","Post-treatment"="#69F5AF","Increase" = "#052BEB", "Decrease" ="#D10A0A", "Missing previous visit"="#A8A7A7","No change"="black"), breaks =c("Increase","Decrease","Missing previous visit"), guide = guide_legend(override.aes =list(linewidth =2, size = 12)))+
  geom_point(position = position_dodge(width = 0.3), size = 5, aes(colour=visit_2))+
     scale_x_discrete(labels =c(
  "Pre-treatment" = "Pre-trt\n(n=27)",  
  "3 Months" = "3 Months\n(n=24)",
  "12 Months" = "12 Months\n(n=22)",
  "Post-treatment" = "Post-trt\n(n=24)"))+
  geom_point(position = position_dodge(width = 0.3), size = 5, colour="black", shape = 21)+
  theme_classic(base_size = 26)+
  scale_y_continuous(limits =c(0,100))+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.position = "bottom",
        legend.title =element_text(size=26))+
labs(title = "", x = "Soap and water treatment", y = "Relative abundance of Staphylococcus %", colour = "Change from previous visit:")
staph_dotplot_water

setwd(f)

#Save as image
png(filename = paste0(y, ".png"),
    width = 2400, height = 4500, res = 300)
staph_dotplot_water
dev.off()

pdf(file = paste0(y, ".pdf"),
    width = 7.63, height = 16.01)
staph_dotplot_water
dev.off()
```

#merge graphs
```{r}
y<-"staph_merged_dotplot"
staph_merged_dotplot <-staph_dotplot_chg +
  staph_dotplot_water +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
    legend.margin = margin(t = 15)
  )

setwd(f)

#Save as image
png(filename = paste0(y, ".png"),
     width = 4400, height = 4500, res = 300)
staph_merged_dotplot
dev.off()

setwd(f)
#Save plot as pdf
pdf(file = paste0(y, ".pdf"),
    width = 14.66, height = 16.01)
staph_merged_dotplot
dev.off()
```

#Water stats
```{r}
meta_data<-df_staph_water


#3 months
Months_3<-subset(meta_data, vcode_prevfootcomp == "3")

print(" 3 month total")
t<-nrow(Months_3)
t

print("Missing previous visit")
nc<-subset(Months_3, Patient_change == "Missing previous visit")
nrow(nc)
nrow(nc)/(t-nrow(nc))*100


print("Increase")
inc<-subset(Months_3, Patient_change == "Increase")
nrow(inc)
nrow(inc)/(t-nrow(nc))*100

summary(inc$change_amount)
mean(inc$change_amount)
sd(inc$change_amount)

print("Decrease")
dc<-subset(Months_3, Patient_change == "Decrease")
nrow(dc)
nrow(dc)/(t-nrow(nc))*100

summary(dc$change_amount)
mean(dc$change_amount)
sd(dc$change_amount)

#12 months
Months_12<-subset(meta_data, vcode_prevfootcomp == "6")

print("12 month total")
t<-nrow(Months_12)
t

print("Missing previous visit")
nc<-subset(Months_12, Patient_change == "Missing previous visit")
nrow(nc)
nrow(nc)/(t-nrow(nc))*100

print("Increase")
inc<-subset(Months_12, Patient_change == "Increase")
nrow(inc)
nrow(inc)/(t-nrow(nc))*100

summary(inc$change_amount)
mean(inc$change_amount)
sd(inc$change_amount)

print("Decrease")
dc<-subset(Months_12, Patient_change == "Decrease")
nrow(dc)
nrow(dc)/(t-nrow(nc))*100

summary(dc$change_amount)
mean(dc$change_amount)
sd(dc$change_amount)

#Post
post<-subset(meta_data, vcode_prevfootcomp == "7")

print("post total")
t<-nrow(post)
t

print("Missing previous visit")
nc<-subset(post, Patient_change == "Missing previous visit")
nrow(nc)
nrow(nc)/(t-nrow(nc))*100

print("Increase")
inc<-subset(post, Patient_change == "Increase")
nrow(inc)
nrow(inc)/(t-nrow(nc))*100

summary(inc$change_amount)
mean(inc$change_amount)
sd(inc$change_amount)

print("Decrease")
dc<-subset(post, Patient_change == "Decrease")
nrow(dc)
nrow(dc)/(t-nrow(nc))*100

summary(dc$change_amount)
mean(dc$change_amount)
sd(dc$change_amount)
```

#Chg stats
```{r}
meta_data<-df_staph_chg

#3 months
Months_3<-subset(meta_data, vcode_prevfootcomp == "3")

print(" 3 month total")
t<-nrow(Months_3)
t

print("Missing previous visit")
nc<-subset(Months_3, Patient_change == "Missing previous visit")
nrow(nc)
nrow(nc)/(t-nrow(nc))*100

print("Increase")
inc<-subset(Months_3, Patient_change == "Increase")
nrow(inc)
nrow(inc)/(t-nrow(nc))*100

summary(inc$change_amount)
mean(inc$change_amount)
sd(inc$change_amount)

print("Decrease")
dc<-subset(Months_3, Patient_change == "Decrease")
nrow(dc)
nrow(dc)/(t-nrow(nc))*100

summary(dc$change_amount)
mean(dc$change_amount)
sd(dc$change_amount)

#12 months
Months_12<-subset(meta_data, vcode_prevfootcomp == "6")

print("12 month total")
t<-nrow(Months_12)
t

print("Missing previous visit")
nc<-subset(Months_12, Patient_change == "Missing previous visit")
nrow(nc)
nrow(nc)/(t-nrow(nc))*100

print("Increase")
inc<-subset(Months_12, Patient_change == "Increase")
nrow(inc)
nrow(inc)/(t-nrow(nc))*100

summary(inc$change_amount)
mean(inc$change_amount)
sd(inc$change_amount)

print("Decrease")
dc<-subset(Months_12, Patient_change == "Decrease")
nrow(dc)
nrow(dc)/(t-nrow(nc))*100

summary(dc$change_amount)
mean(dc$change_amount)
sd(dc$change_amount)

#Post
post<-subset(meta_data, vcode_prevfootcomp == "7")

print("post total")
t<-nrow(post)
t

print("Missing previous visit")
nc<-subset(post, Patient_change == "Missing previous visit")
nrow(nc)
nrow(nc)/(t-nrow(nc))*100

print("Increase")
inc<-subset(post, Patient_change == "Increase")
nrow(inc)
nrow(inc)/(t-nrow(nc))*100

summary(inc$change_amount)
mean(inc$change_amount)
sd(inc$change_amount)

print("Decrease")
dc<-subset(post, Patient_change == "Decrease")
nrow(dc)
nrow(dc)/(t-nrow(nc))*100

summary(dc$change_amount)
mean(dc$change_amount)
sd(dc$change_amount)

```


#Merge figure 4 plots
```{r}
x <-"figure_4_merged"

combined_relative_abundance_2 <- combined_relative_abundance
staph_merged_dotplot_2 <- staph_merged_dotplot 


figure_4<- (combined_relative_abundance_2 | staph_merged_dotplot_2)+  
  plot_layout(widths = c(1,1.1))

figure_4

  
#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 31.333, height =  18) 
figure_4
dev.off()

#Save as image
png(filename = paste0(x,".png"), width =  9400 , height = 5400, res = 300)
figure_4
dev.off()
```

