---
title: "Maryland_Supplemental_Figures"
output: html_document
date: "2025-12-31"
note: must run script "Maryland_Final_Figures.rmd" first.
---

#Set working directory
```{r setup, include=TRUE}
knitr::opts_knit$set(root.dir = "[input pathway to directory with data]")

#Directory for saving figures
f<-"[input pathway to where figure will be saved]"
```

#Activate packages
```{r}
library("phyloseq")
library("ggplot2")
library("patchwork")
library("ggpubr")
library("ALDEx2")
library("phyloseq")
library("dplyr")
library("tidyverse")
library ("BiocManager")
library("devtools")
library("httr")
library("jsonlite") 
library("reticulate")
library("qiime2R")
library("vegan")
library("decontam")
```

#Supplemental Figure 1:ITS Figures Part 1

```{r}
water_rare_genus_ITS<-subset_samples(ps_rare_genus_ITS, treatgroup_2 == "Soap and water")
chg_rare_genus_ITS<-subset_samples(ps_rare_genus_ITS, treatgroup_2 == "Chlorhexidine")
water_rare_genus_ITS
chg_rare_genus_ITS
```

#Figure 1A:Bray-Curtis Change Overtime

#CHG
#Prepare data
```{r}
#Ps object with exact counts
ps_bds<-chg_rare_genus_ITS

sample_data(ps_bds)$visit_2 <- fct_recode(sample_data(ps_bds)$vcode_prevfootcomp, 
                                        "Pre-trt" = "1", 
                                        "3 Months" = "3", 
                                        "12 Months" = "6",
                                        "Post-trt" = "7")
#set testing parameter
x="visit_2"

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot

print("permanova")
permanova_result <- vegan::adonis2(bray_dist ~ visit_2, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)

bray_visit_chg_ITS<-ps_bds
```


#Graph
```{r}
y<-"Bray_Curtis_visit_chg_ITS"

# Run NMDS 
ordination_result <- ordinate(bray_visit_chg_ITS, method='NMDS', distance='bray', k=4)

#Check stress levels if >0.2 increase k. ideally <0.1
ordination_result$stress

# Plot Axis 1 and Axis 2
bray_visit_chg_plot_ITS<-plot_ordination(bray_visit_chg_ITS, ordination_result, axes=c(1,2), color=x)+
geom_point(size = 6) +
  geom_point(
    size = 6,
    shape = 21,        
    color = "black",   
    stroke = 0.3) +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
))) +
  theme_classic(base_size = 20)+
  theme_classic()+
  theme_bw()+
     theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
       legend.position = "bottom") +
  scale_color_manual(values = c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF")) +
  labs( title = "", color = "" )

bray_visit_chg_plot_ITS
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 8) # Adjust dimensions as needed
bray_visit_chg_plot_ITS
dev.off()

#Save plot as pdf
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 2000, res = 300)
bray_visit_chg_plot_ITS
dev.off()
```


#Water

#Prepare data
```{r}
#Ps object with exact counts
ps_bds<-water_rare_genus_ITS

#reformat label
sample_data(ps_bds)$visit_2 <- fct_recode(sample_data(ps_bds)$vcode_prevfootcomp, 
                                        "Pre-trt" = "1", 
                                        "3 Months" = "3", 
                                        "12 Months" = "6",
                                        "Post-trt" = "7")

#set testing parameter
x="visit_2"

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot

print("permanova")
permanova_result <- vegan::adonis2(bray_dist ~ visit_2, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)

bray_visit_water_ITS<-ps_bds
```

#Graph
```{r}
y<-"Bray_Curtis_visit_water_ITS"


# Run NMDS 
ordination_result <- ordinate(bray_visit_water_ITS, method='NMDS', distance='bray', k=4)

#Check stress levels if >0.2 increase k. ideally <0.1
ordination_result$stress

# Plot Axis 1 and Axis 2
bray_visit_water_plot_ITS<-plot_ordination(bray_visit_water_ITS, ordination_result, axes=c(1,2), color=x)+
geom_point(size = 6) +
  geom_point(
    size = 6,
    shape = 21,        
    color = "black",   
    stroke = 0.3) +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
))) +
  theme_classic()+
  theme_bw()+
     theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
       legend.position = "bottom") +
  scale_color_manual(values = c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF")) +
  labs( title = "", color = "" )

bray_visit_water_plot_ITS
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 8) # Adjust dimensions as needed
bray_visit_water_plot_ITS
dev.off()

#Save plot as pdf
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 2000, res = 300)
bray_visit_water_plot_ITS
dev.off()
```

#Combine Graphs
```{r}
x<-"combined_BC_visit_ITS"

# p1 has legend removed
p1 <- bray_visit_chg_plot_ITS + 
  theme(legend.position = "none") + 
  labs(title = "Chlorhexidine treatment") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.margin = margin(b = 30))

# p2 keeps legend
p2 <- bray_visit_water_plot_ITS + 
   theme(legend.position = "bottom",legend.margin = margin(t = 15), plot.margin = margin(t = 30)
  )+
  labs(title = "Soap and water treatment") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

combined_BC_visit_ITS <- p1 / p2 

combined_BC_visit_ITS
```

```{r}
#Save plot as pdf

setwd(f)
pdf(file = paste0(x,".pdf"), width = 10.333, height =  7.5) 
combined_BC_visit
dev.off()

#Save as image
png(filename = paste0(x,".png"), width =  3000 , height = 5400, res = 300)
combined_BC_visit
dev.off()
```




#Supplemental Figure 1B:Bray curtis plots

#Chg
```{r}
##Prepare data

#transform to relative abundance 
rel_abundance = transform_sample_counts(chg_rare_genus_ITS, function(x) x*100/ sum(x))

#Ps object with relative abundance
ps_bds <- rel_abundance 

#calculate bray-distance
bray_dist <- distance(ps_bds, method="bray") %>%
  as.matrix() %>%
  as.data.frame()

#Structure table into rows (FIXED: create bray_long)
bray_long <- rownames_to_column(bray_dist, "Sample_1") %>%
  pivot_longer(
    -Sample_1,
    names_to = "Sample_2",
    values_to = "Bray_Curtis",
    values_transform = list(Bray_Curtis = as.numeric)
  )

#format metadata
meta <- as(sample_data(ps_bds), "data.frame") %>%
  rownames_to_column("Sample_ID")

#FIXED: use bray_long
merged_df <- bray_long %>%
  left_join(meta, by = c("Sample_1" = "Sample_ID"))

#Identify visit 1 sample IDs
v1_list <- merged_df %>%
  filter(vcode_prevfootcomp == "1") %>%
  pull(Sample_1) %>%
  unique()

#Isolate visit 1 comparisons
vist_1 <- merged_df %>%
  filter(Sample_2 %in% v1_list)

#Remove duplicate sample pairs
vist_1 <- vist_1 %>%
  mutate(
    pair_min = pmin(Sample_1, Sample_2),
    pair_max = pmax(Sample_1, Sample_2)
  ) %>%
  distinct(pair_min, pair_max, .keep_all = TRUE)

#Remove same-sample comparisons
vist_1 <- vist_1 %>%
  filter(Sample_1 != Sample_2)

#Save output
chg_v1_BC <- vist_1
chg_v1_BC
```

#Plot figure
```{r}
ps <- chg_v1_BC

#Order factors correctly
ps$visit_2 <- factor(ps$visit_2, levels = c("Pre-treatment","3 Months","12 Months", "Post-treatment"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment","3 Months"),c("Pre-treatment","12 Months"),c("Pre-treatment","Post-treatment"))

stats<-compare_means( Bray_Curtis ~ visit_2,
              data =  chg_v1_BC,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

y<-"CHG_BC_Boxplots_ITS"

BC_Boxplot_chg_ITS<-ggplot(ps, aes(x=visit_2,y=Bray_Curtis, fill= visit_2)) +
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)+
  theme_classic(base_size = 20)+
  theme_classic(base_size = 20)+
  scale_x_discrete(labels =c(
  "Pre-treatment" = "Pre-trt\n(n=26)",  
  "3 Months" = "3 Months\n(n=25)",
  "12 Months" = "12 Months\n(n=17)",
  "Post-treatment" = "Post-trt\n(n=28)"))+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
        legend.position = "none", 
       strip.text = element_blank())+
  scale_color_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     size = 10,
                     comparisons = comparisons)+
  labs(color = "Visit", x = "Chlorhexidine treatment", y = "Bray-Curtis Dissimilarity")

BC_Boxplot_chg_ITS
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y,".pdf"), width = 6.67, height = 8.1) # Adjust dimensions as needed
BC_Boxplot_chg_ITS
dev.off()

#Save as image
png(filename = paste0(y,".png"), width = 2000, height = 2500, res = 300)
BC_Boxplot_chg_ITS
dev.off()
```

#Average distance
```{r}
interest_group<-chg_v1_BC

summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean_dev = mean(Bray_Curtis, na.rm = TRUE),
    sd_dev = sd(Bray_Curtis, na.rm = TRUE),
    se = sd_dev / sqrt(n),
    ci95_low  = mean_dev - qt(0.975, df = n-1) * se,
    ci95_high = mean_dev + qt(0.975, df = n-1) * se
  )

summary_stats
```


#Water
```{r}
#Prepare data

#transform to relative abundance 
rel_abundance = transform_sample_counts(water_rare_genus_ITS, function(x) x*100/ sum(x))

#Ps object with relative abundance
ps_bds <- rel_abundance 

#calculate bray-distance
bray_dist <- distance(ps_bds, method="bray") %>%
  as.matrix() %>%
  as.data.frame()

#Structure table into rows (FIXED: create bray_long)
bray_long <- rownames_to_column(bray_dist, "Sample_1") %>%
  pivot_longer(
    -Sample_1,
    names_to = "Sample_2",
    values_to = "Bray_Curtis",
    values_transform = list(Bray_Curtis = as.numeric)
  )

#format metadata
meta <- as(sample_data(ps_bds), "data.frame") %>%
  rownames_to_column("Sample_ID")

#FIXED: use bray_long
merged_df <- bray_long %>%
  left_join(meta, by = c("Sample_1" = "Sample_ID"))

#Identify visit 1 sample IDs
v1_list <- merged_df %>%
  filter(vcode_prevfootcomp == "1") %>%
  pull(Sample_1) %>%
  unique()

#Isolate visit 1 comparisons
vist_1 <- merged_df %>%
  filter(Sample_2 %in% v1_list)

#Remove duplicate sample pairs
vist_1 <- vist_1 %>%
  mutate(
    pair_min = pmin(Sample_1, Sample_2),
    pair_max = pmax(Sample_1, Sample_2)
  ) %>%
  distinct(pair_min, pair_max, .keep_all = TRUE)

#Remove same-sample comparisons
vist_1 <- vist_1 %>%
  filter(Sample_1 != Sample_2)

#Save output
water_v1_BC <- vist_1
water_v1_BC
```

#Plot figure
```{r}
ps <- water_v1_BC


#Order factors correctly
ps$visit_2 <- factor(ps$visit_2, levels = c("Pre-treatment","3 Months","12 Months", "Post-treatment"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment","3 Months"),c("Pre-treatment","12 Months"),c("Pre-treatment","Post-treatment"))

stats<-compare_means( Bray_Curtis ~ visit_2,
              data =  water_v1_BC,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

y<-"water_BC_Boxplots_ITS"

BC_Boxplot_water_ITS<-ggplot(ps, aes(x=visit_2,y=Bray_Curtis, fill= visit_2)) +
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)+
  theme_classic(base_size = 20)+
  theme_classic(base_size = 20)+
   scale_x_discrete(labels =c(
  "Pre-treatment" = "Pre-trt\n(n=26)",  
  "3 Months" = "3 Months\n(n=20)",
  "12 Months" = "12 Months\n(n=14)",
  "Post-treatment" = "Post-trt\n(n=17)"))+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
        legend.position = "none", 
       strip.text = element_blank())+
  scale_color_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     size = 10,
                     comparisons = comparisons)+
  labs(color = "Visit", x = "Soap and water treatment", y = "Bray-Curtis Dissimilarity")

BC_Boxplot_water_ITS
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y,".pdf"), width = 6.67, height = 8.1) # Adjust dimensions as needed
BC_Boxplot_water_ITS
dev.off()

#Save as image
png(filename = paste0(y,".png"), width = 2000, height = 2500, res = 300)
BC_Boxplot_water_ITS
dev.off()
```

#Average distance
```{r}
interest_group<-water_v1_BC

summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean_dev = mean(Bray_Curtis, na.rm = TRUE),
    sd_dev = sd(Bray_Curtis, na.rm = TRUE),
    se = sd_dev / sqrt(n),
    ci95_low  = mean_dev - qt(0.975, df = n-1) * se,
    ci95_high = mean_dev + qt(0.975, df = n-1) * se
  )

summary_stats
```


#Combine Graphs
```{r}
x<-"combined_betadiversity_boxplots_ITS"

# p1 has legend removed
p1 <- BC_Boxplot_chg_ITS + 
  theme(legend.position = "none") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.margin = margin(b = 30))

# p2 keeps legend
p2 <- BC_Boxplot_water_ITS + 
   theme(legend.position = "none",legend.margin = margin(t = 15), plot.margin = margin(t = 30)
  )+
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

combined_betadiversity_boxplots_ITS <- p1 / p2 

combined_betadiversity_boxplots_ITS

```

```{r}
#Save plot as pdf

setwd(f)
pdf(file = paste0(x,".pdf"), width = 10.333, height =  7.7) 
combined_betadiversity_boxplots_ITS
dev.off()

#Save as image
png(filename = paste0(x,".png"), width =  3000 , height = 5700, res = 300)
combined_betadiversity_boxplots_ITS
dev.off()

```

#SupplementalFigure 1C:Change during treatment ITS

#CHG
#Prepare data
```{r}
#Ps object with exact counts
ps_bds<-chg_rare_genus_ITS

#set testing parameter
x="treatment_active"

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot

print("permanova")
permanova_result <- vegan::adonis2(bray_dist ~ treatment_active, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)

bray_active_treat_chg_ITS<-ps_bds
```


#Graph
```{r}
y<-"Bray_Curtis_no_active_treatment_chg_ITS"

# Run NMDS 
ordination_result <- ordinate(bray_active_treat_chg_ITS, method='NMDS', distance='bray', k=4)

#Check stress levels if >0.2 increase k. ideally <0.1
ordination_result$stress

# Plot Axis 1 and Axis 2
bray_active_treat_chg_plot_ITS<-plot_ordination(bray_active_treat_chg_ITS, ordination_result, axes=c(1,2), color=x)+
geom_point(size = 6) +
  geom_point(
    size = 6,
    shape = 21,        
    color = "black",   
    stroke = 0.3) +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#F54E4E","#8953CF"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
))) +
  theme_classic(base_size = 20)+
  theme_classic()+
  theme_bw()+
     theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
       legend.position = "bottom") +
  scale_color_manual(
  values = c(
    "Active treatment (3 months + 12 months)" = "#F54E4E",
    "No active treatment (pre-trt + post-trt)" = "#8953CF"
  ),
  labels = c(
    "Active treatment (3 months + 12 months)" = "Active trt (3 & 12 months)",
    "No active treatment (pre-trt + post-trt)" = "No active trt (pre & post-trt)"
  )
)+
  labs( title = "", color = "" )

bray_active_treat_chg_plot_ITS
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 8) # Adjust dimensions as needed
bray_active_treat_chg_plot_ITS
dev.off()

#Save plot as pdf
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 2000, res = 300)
bray_active_treat_chg_plot_ITS
dev.off()
```


#Water

#Prepare data
```{r}
#Ps object with exact counts
ps_bds<-water_rare_genus_ITS

#set testing parameter
x="treatment_active"

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot

print("permanova")
permanova_result <- vegan::adonis2(bray_dist ~ treatment_active, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)

bray_active_treat_water_ITS<-ps_bds
```

#Graph
```{r}
y<-"Bray_Curtis_no_active_treatment_water_ITS"

# Run NMDS 
ordination_result <- ordinate(bray_active_treat_water_ITS, method='NMDS', distance='bray', k=4)

#Check stress levels if >0.2 increase k. ideally <0.1
ordination_result$stress

# Plot Axis 1 and Axis 2
bray_active_treat_water_plot_ITS<-plot_ordination(bray_active_treat_water_ITS, ordination_result, axes=c(1,2), color=x)+
geom_point(size = 6) +
  geom_point(
    size = 6,
    shape = 21,        
    color = "black",   
    stroke = 0.3) +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#F54E4E","#8953CF"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
))) +
  theme_classic()+
  theme_bw()+
     theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
       legend.position = "bottom") +
   scale_color_manual(
  values = c(
    "Active treatment (3 months + 12 months)" = "#F54E4E",
    "No active treatment (pre-trt + post-trt)" = "#8953CF"
  ),
  labels = c(
    "Active treatment (3 months + 12 months)" = "Active trt (3 & 12 months)",
    "No active treatment (pre-trt + post-trt)" = "No active trt (pre & post-trt)"
  )
)+
  labs( title = "", color = "" )

bray_active_treat_water_plot_ITS
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 8) # Adjust dimensions as needed
bray_active_treat_water_plot_ITS
dev.off()

#Save plot as pdf
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 2000, res = 300)
bray_active_treat_water_plot_ITS
dev.off()
```


#Combine Graphs
```{r}
x<-"combined_active_treatment"

# p1 has legend removed
p1 <- bray_active_treat_chg_plot_ITS + 
  labs(title = "Chlorhexidine treatment") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.margin = margin(b = 30))
p1 
# p2 keeps legend
p2 <- bray_active_treat_water_plot_ITS + 
  labs(title = "Soap and water treatment") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

combined_active_treatment_ITS <- (p1 / p2) +
   plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

combined_active_treatment_ITS
```

```{r}
#Save plot as pdf

setwd(f)
pdf(file = paste0(x,".pdf"), width = 10.333, height =  7.5) 
combined_active_treatment_ITS
dev.off()

#Save as image
png(filename = paste0(x,".png"), width =  3000 , height = 5400, res = 300)
combined_active_treatment_ITS
dev.off()

```


#Merge supplemental figure 1 plots
```{r}
x <-"Supplemental_figure_1_merged"

combined_BC_visit_ITS_2  <- combined_BC_visit_ITS + theme(plot.margin = margin(b = 0,
                                                                              t = 50,
                                                                              r = 30,
                                                                              l =50))


combined_betadiversity_boxplots_2 <- combined_betadiversity_boxplots_ITS + theme(plot.margin = margin(b = 0,
                                                                              t = 50,
                                                                              r = 30,
                                                                              l =50))




combined_active_treatment_2 <- combined_active_treatment_ITS + theme(plot.margin = margin(b = 30,
                                            t = 30,
                                            l = 50,
                                            r = 50)) 






figure_1_ITS <- (combined_BC_visit_ITS_2 |
  combined_betadiversity_boxplots_2 | 
  combined_active_treatment_2 ) + 
  plot_layout(widths = c(1,1,1.25))


#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 31.333, height =  18) 
figure_1_ITS
dev.off()
#Save as image
png(filename = paste0(x,".png"), width =  9400 , height = 5400, res = 300)
figure_1_ITS
dev.off()
```



#Supplemental Figure 2:ITS Figures Part 2

#Supplemental Figure 2A: ITS Relative abundance 

#Prepare data
```{r}
#Convert data structure 
rel_abundance_ITS<-transform_sample_counts(ps_nr_genus_ITS, function(x) x*100/sum(x))
df<-psmelt(rel_abundance_ITS)

#Isolate treatment groups
water_df<-subset(df, treatgroup == "0")
chg_df<-subset(df, treatgroup == "1")

#Isolate top abundant genera 
top_genera_w <- water_df %>%
  group_by(Genus) %>%
  summarize(TotalAbundance = sum(Abundance, na.rm = TRUE)) %>%
  arrange(desc(TotalAbundance)) %>%
  slice_head(n = 15)  # Select the top 15 genera

top_genus_names_w<-top_genera_w$Genus

water_df$Genus<-ifelse(water_df$Genus %in% top_genus_names_w, water_df$Genus, "Other")


top_genera_c <- chg_df %>%
  group_by(Genus) %>%
  summarize(TotalAbundance = sum(Abundance, na.rm = TRUE)) %>%
  arrange(desc(TotalAbundance)) %>%
  slice_head(n = 15)  # Select the top 15 genera

top_genus_names_c<-top_genera_c$Genus
chg_df$Genus<-ifelse(chg_df$Genus %in% top_genus_names_c, chg_df$Genus, "Other")

#Check
chg_df
water_df

top_genera_c
top_genus_names_c
```


#Bubble Plot
```{r}
y<-"ITS_Rel_chg"

ps <-chg_df
order<-c("Other",top_genus_names_c)
ps$Genus<-factor(ps$Genus, levels = order)


ITS_Rel_chg<-ggplot(ps, aes(x=Patient_id, y=Genus, size = Abundance)) +
  geom_point(alpha=0.7)+
  facet_grid(~visit_2, scales = "free_x", space = "free_x", labeller = labeller(
  visit_2 = c(
    "Pre-treatment"  = "Pre-trt\n(n=26)",
    "3 Months"       = "3 Months\n(n=25)",
    "12 Months"      = "12 Months\n(n=17)",
    "Post-treatment" = "Post-trt\n(n=28)"
)))+
 theme(strip.text = element_text(size=20),
       panel.background = element_rect(fill="#fafafa"),
   strip.background = element_blank(), 
        axis.text.y = element_text(size=20,face ="italic"),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1,face = "bold"),
        axis.text.x.bottom =  element_blank(),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26))+
  labs(color = "visit", size = "Relative Abundance %", x ="Chlorhexidine treatment", y="Top 15 Genera")

ITS_Rel_chg

#Save plot as pdf
setwd(f)
pdf(file = paste0(y, ".pdf"), width = 40, height = 6) 
print(ITS_Rel_chg)
dev.off()


#Save as image
png(filename = paste0(y, ".png"), width = 6000, height = 2000, res = 300)
print(ITS_Rel_chg) 
dev.off()
```

```{r}
y<-"ITS_Rel_water"

ps <-water_df
order<-c("Other",top_genus_names_w)
ps$Genus<-factor(ps$Genus, levels = order)


ITS_Rel_water<-ggplot(ps, aes(x=Patient_id, y=Genus, size = Abundance)) +
  geom_point(alpha=0.7)+
  facet_grid(~visit_2, scales = "free_x", space = "free_x", labeller = labeller(
  visit_2 = c(
    "Pre-treatment"  = "Pre-trt\n(n=26)",
    "3 Months"       = "3 Months\n(n=20)",
    "12 Months"      = "12 Months\n(n=14)",
    "Post-treatment" = "Post-trt\n(n=17)"
  )))+
theme(strip.text = element_text(size=20),
       panel.background = element_rect(fill="#fafafa"),
   strip.background = element_blank(), 
        axis.text.y = element_text(size=20,face ="italic"),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1,face = "bold"),
        axis.text.x.bottom =  element_blank(),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26))+
  labs(color = "visit", size = "Relative Abundance %", x ="Soap and water treatment", y="Top 15 Genera")

ITS_Rel_water

#Save plot as pdf
setwd(f)
pdf(file = paste0(y, ".pdf"), width = 40, height = 6) # Adjust dimensions as needed
print(ITS_Rel_water) 
dev.off()


#Save as image
png(filename = paste0(y, ".png"), width = 6000, height = 2000, res = 300)
print(ITS_Rel_water) 
dev.off()
```




#Supplemental Figure 2B: Alpha diversity

#Water
```{r}
library(ggplot2)
library(ggstatsplot)
library(ggpubr)

#Pick_phyloseq group
water_rare_genus_ITS<-subset_samples(ps_rare_genus_ITS, treatgroup == "0")
ps <- water_rare_genus_ITS

#Order factors correctly
sample_data(ps)$visit_2 <- factor(sample_data(ps)$visit_2, levels = c("Pre-treatment","3 Months","12 Months", "Post-treatment"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment","3 Months"),c("Pre-treatment","12 Months"),c("Pre-treatment","Post-treatment"))


#Title of graph *CHANGE EACH TIME
x <-"Alpha_Diversity_shannon_change_overtime_water_samples_ITS"

#Check if adjusted p-values align with results determined later
rich <- estimate_richness(ps, measures = "Shannon")
df <- cbind(as.data.frame(sample_data(ps)), rich)

stats<-compare_means(Shannon ~ visit_2,
              data = df,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

#Include n-values
meta<-sample_data(ps)

#Generate plot 
water_alpha_diversity_shannon_ITS<- plot_richness(ps, x="visit_2", title = NULL, measures=c("Shannon"), color = "visit_2")+
  geom_jitter(
  aes(fill = visit_2),
  shape = 21,
  color = "black",
  stroke = 0.3,
  width = 0.10,
  size = 3,
  alpha = 1)+
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)+
 scale_y_continuous(limits = c(0,4.2))+
   scale_x_discrete(labels =c(
  "Pre-treatment" = "Pre-trt\n(n=26)",  
  "3 Months" = "3 Months\n(n=20)",
  "12 Months" = "12 Months\n(n=14)",
  "Post-treatment" = "Post-trt\n(n=17)"))+
  theme_classic(base_size = 20)+
 theme(strip.background = , 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
        legend.position = "none",
        strip.text = element_blank())+
  scale_color_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     position = "top",
                     size = 12,
                     label.y = c(2.5,3.2,3.9),
                     comparisons = comparisons)+
  labs(color = "Visit", x = "Soap and water treatment", y = "Shannon diversity index")

#remove defult points
water_alpha_diversity_shannon_ITS$layers <- water_alpha_diversity_shannon_ITS$layers[-1]

water_alpha_diversity_shannon_ITS


#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 6.67, height = 8.1) 
print(water_alpha_diversity_shannon_ITS) 
dev.off()

#Save as image
png(filename = paste0(x, ".png"), width = 2000, height = 2500, res = 300)
print(water_alpha_diversity_shannon_ITS) 
dev.off()
```

#Average diversity
```{r}
ps<-water_rare_genus_ITS
rich <- estimate_richness(ps, measures = "Shannon")
interest_group <- cbind(as.data.frame(sample_data(ps)), rich)


summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean = mean(Shannon, na.rm = TRUE),
    sd = sd(Shannon, na.rm = TRUE),
    se = sd / sqrt(n),
    ci95_low  = mean - qt(0.975, df = n-1) * se,
    ci95_high = mean + qt(0.975, df = n-1) * se
  )


summary_stats
```


#CHG
```{r}
library(ggplot2)
library(ggstatsplot)
library(ggpubr)

#Pick_phyloseq group
chg_rare_genus_ITS<-subset_samples(ps_rare_genus_ITS, treatgroup == "1")
ps <- chg_rare_genus_ITS

#Order factors correctly
sample_data(ps)$visit_2 <- factor(sample_data(ps)$visit_2, levels = c("Pre-treatment","3 Months","12 Months", "Post-treatment"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment","3 Months"),c("Pre-treatment","12 Months"),c("Pre-treatment","Post-treatment"))


#Title of graph *CHANGE EACH TIME
x <-"Alpha_Diversity_shannon_change_overtime_chg_samples_ITS"

#Check if adjusted p-values align with results determined later
rich <- estimate_richness(ps, measures = "Shannon")
df <- cbind(as.data.frame(sample_data(ps)), rich)

stats<-compare_means(Shannon ~ visit_2,
              data = df,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

#Include n-values
meta<-sample_data(ps)

#Generate plot 
chg_alpha_diversity_shannon_ITS<- plot_richness(ps, x="visit_2", title = NULL, measures=c("Shannon"), color = "visit_2")+
  geom_jitter(
  aes(fill = visit_2),
  shape = 21,
  color = "black",
  stroke = 0.3,
  width = 0.10,
  size = 3,
  alpha = 1)+
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)+
 scale_y_continuous(limits = c(0,4.2))+
   scale_x_discrete(labels =c("Pre-treatment" = "Pre-trt\n(n=26)",  
  "3 Months" = "3 Months\n(n=25)",
  "12 Months" = "12 Months\n(n=17)",
  "Post-treatment" = "Post-trt\n(n=28)"))+
  theme_classic(base_size = 20)+
 theme(strip.background = , 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
        legend.position = "none",
        strip.text = element_blank())+
  scale_color_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     position = "top",
                     size = 12,
                     label.y = c(2.5,3.2,3.9),
                     comparisons = comparisons)+
  labs(color = "Visit", x = "Chlorhexidine treatment", y = "Shannon diversity index")

#remove defult points
chg_alpha_diversity_shannon_ITS$layers <- chg_alpha_diversity_shannon_ITS$layers[-1]

chg_alpha_diversity_shannon_ITS


#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 6.67, height = 8.1) 
print(chg_alpha_diversity_shannon_ITS) 
dev.off()

#Save as image
png(filename = paste0(x, ".png"), width = 2000, height = 2500, res = 300)
print(chg_alpha_diversity_shannon_ITS) 
dev.off()
```

#Average diversity
```{r}
ps<-chg_rare_genus_ITS
rich <- estimate_richness(ps, measures = "Shannon")
interest_group <- cbind(as.data.frame(sample_data(ps)), rich)


summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean = mean(Shannon, na.rm = TRUE),
    sd = sd(Shannon, na.rm = TRUE),
    se = sd / sqrt(n),
    ci95_low  = mean - qt(0.975, df = n-1) * se,
    ci95_high = mean + qt(0.975, df = n-1) * se
  )

summary_stats
```

#Merge Supplemental figure 2
```{r}
x <- "Supplemental_figure_2"

Rel_chg <- ITS_Rel_chg + 
  theme(plot.margin = margin(b = 30, t = 30, l = 30, r = 50),
        legend.position = "none")

Rel_water <- ITS_Rel_water + 
  theme(plot.margin = margin(b = 30, t = 30, l = 30, r = 50),
        legend.position = "bottom")

Rel_ITS <- (Rel_chg / Rel_water) &
  theme(plot.margin = margin(b = 30, t = 30, l = 30, r = 50),
        legend.margin = margin(t = 15)) +
  plot_layout(heights = c(1,1))

alpha_ITS <- (chg_alpha_diversity_shannon_ITS / water_alpha_diversity_shannon_ITS) &
  theme(plot.margin = margin(b = 30, t = 30, l = 30, r = 50))

figure_2_s <- (Rel_ITS | alpha_ITS) +
  plot_layout(widths = c(3,1.5))


#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 31.333, height =  18) 
figure_2_s
dev.off()
#Save as image
png(filename = paste0(x,".png"), width =  9400 , height = 5400, res = 300)
figure_2_s
dev.off()
```






#Supplemental Figure 3A: Bioburden 

#Graph of CHG bioburden box plot seperate 
```{r}

#Order factors correctly
grouped_chg_dpcr$visit_2 <- factor(grouped_chg_dpcr$visit_2, levels = c("Pre-treatment (n=45)","3 Months (n=38)","12 Months (n=31)", "Post-treatment (n=34)"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment (n=45)","3 Months (n=38)"),c("Pre-treatment (n=45)","12 Months (n=31)"),c("Pre-treatment (n=45)","Post-treatment (n=34)"))

#Check if adjusted p-values align with results determined later
stats<-compare_means( log_Concentration ~ visit_2,
              data = grouped_chg_dpcr,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

y <-"bioburden_chg_boxplot"

#Visualization 

library(ggplot2)

bioburden_boxplot_chg<-ggplot(grouped_chg_dpcr, aes(x=visit_2, y=log_Concentration, fill=visit_2))+
  geom_jitter(
  aes(fill = visit_2),
  shape = 21,
  color = "black",
  stroke = 0.3,
  width = 0.15,
  size = 2,
  alpha = 0.5)+
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)  +
  scale_x_discrete(labels =c(
  "Pre-treatment (n=45)" = "Pre-trt\n(n=45)",  
  "3 Months (n=38)" = " 3 Months\n(n=38)",
  "12 Months (n=31)" = "12 Months\n(n=31)",
  "Post-treatment (n=34)" = "Post-trt\n(n=34)"))+
  theme_classic(base_size = 20)+
  scale_y_continuous(
  limits = c(0.5, 6.3),
  breaks = seq(0, 5, 1),
  expand = expansion(mult = c(0, 0))
)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.position = "none",
        legend.title =element_text(size=26))+
labs(title = "", x = "Chlorhexidine treatment", y = expression(bold("log"[10]* " Concentration")), colour = "Change from previous visit:")+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     label.y = c(4.3,5.0,5.7),
                     size = 12,
                     comparisons = comparisons)
  
bioburden_boxplot_chg
```

```{r}
setwd(f)

#Save as image
png(filename = paste0(y,".png"), width = 3000, height = 2000, res = 300)
bioburden_boxplot_chg # Replace with your actual graph object
dev.off()

#Save plot as pdf
pdf(file = paste0(y,".pdf"), width = 14, height = 7) # Adjust dimensions as needed
bioburden_boxplot_chg # Replace with your plot object
dev.off()
```

#Average concentration
```{r}
interest_group<-grouped_chg_dpcr

summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean_dev = mean(log_Concentration, na.rm = TRUE),
    sd_dev = sd(log_Concentration, na.rm = TRUE),
    se = sd_dev / sqrt(n),
    ci95_low  = mean_dev - qt(0.975, df = n-1) * se,
    ci95_high = mean_dev + qt(0.975, df = n-1) * se
  )

summary_stats
```



#Graph of water bioburden box plot seperate 
```{r}
#Order factors correctly
grouped_water_dpcr$visit_2 <- factor(grouped_water_dpcr$visit_2, levels = c("Pre-treatment (n=38)","3 Months (n=32)","12 Months (n=28)","Post-treatment (n=31)"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment (n=38)","3 Months (n=32)"),c("Pre-treatment (n=38)","12 Months (n=28)"),c("Pre-treatment (n=38)","Post-treatment (n=31)"))

#Check if adjusted p-values align with results determined later
stats<-compare_means(log_Concentration ~ visit_2,
              data = grouped_water_dpcr,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

y <-"bioburden_water_boxplot"

#Visualization 

library(ggplot2)

bioburden_boxplot_water<-ggplot(grouped_water_dpcr, aes(x=visit_2, y=log_Concentration, fill=visit_2))+
  geom_jitter(
  aes(fill = visit_2),
  shape = 21,
  color = "black",
  stroke = 0.3,
  width = 0.15,
  size = 2,
  alpha = 0.5)+
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)  +
  scale_x_discrete(labels =c(
  "Pre-treatment (n=38)" = "Pre-trt\n(n=38)",  
  "3 Months (n=32)" = " 3 Months\n(n=32)",
  "12 Months (n=28)" = "12 Months\n(n=28)",
  "Post-treatment (n=31)" = "Post-trt\n(n=31)"))+
  theme_classic(base_size = 20)+
  scale_y_continuous(
  limits = c(0.5, 6.3),
  breaks = seq(0, 5, 1),
  expand = expansion(mult = c(0, 0))
)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.position = "none",
        legend.title =element_text(size=26))+
labs(title = "", x = "Soap and water treatment", y = expression(bold("log"[10]* " Concentration")), colour = "Change from previous visit:")+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     label.y = c(4.3,5.0,5.7),
                     size = 12,
                     comparisons = comparisons)
  
bioburden_boxplot_water
```

```{r}
setwd(f)

#Save as image
png(filename = paste0(y,".png"), width = 3000, height = 2000, res = 300)
bioburden_boxplot_water # Replace with your actual graph object
dev.off()

#Save plot as pdf
pdf(file = paste0(y,".pdf"), width = 14, height = 7) # Adjust dimensions as needed
bioburden_boxplot_water # Replace with your plot object
dev.off()
```

#Average concentration
```{r}
interest_group<-grouped_water_dpcr

summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean_dev = mean(log_Concentration, na.rm = TRUE),
    sd_dev = sd(log_Concentration, na.rm = TRUE),
    se = sd_dev / sqrt(n),
    ci95_low  = mean_dev - qt(0.975, df = n-1) * se,
    ci95_high = mean_dev + qt(0.975, df = n-1) * se
  )

summary_stats
```



#Supplemental Figure 3B:Bray-Curtis Change Overtime

#CHG
#Prepare data
```{r}
#Ps object with exact counts
ps_bds<-chg_rare_ps

sample_data(ps_bds)$visit_2 <- fct_recode(sample_data(ps_bds)$vcode_prevfootcomp, 
                                        "Pre-trt" = "1", 
                                        "3 Months" = "3", 
                                        "12 Months" = "6",
                                        "Post-trt" = "7")
#set testing parameter
x="visit_2"

#Import tree
phy_tree(ps_bds)<-qza_to_phyloseq(tree="/Users/michellebode/maryland/16S/processed_data/tree/rooted_tree.qza")
ps_bds

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot

print("permanova")
permanova_result <- vegan::adonis2(bray_dist ~ visit_2, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)

bray_visit_chg<-ps_bds
```


#Graph
```{r}
y<-"Bray_Curtis_visit_chg"

# Run NMDS 
ordination_result <- ordinate(bray_visit_chg, method='NMDS', distance='bray', k=4)

#Check stress levels if >0.2 increase k. ideally <0.1
ordination_result$stress

# Plot Axis 1 and Axis 2
bray_visit_chg_plot<-plot_ordination(bray_visit_chg, ordination_result, axes=c(1,2), color=x)+
geom_point(size = 6) +
  geom_point(
    size = 6,
    shape = 21,        
    color = "black",   
    stroke = 0.3) +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
))) +
  theme_classic(base_size = 20)+
  theme_classic()+
  theme_bw()+
     theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
       legend.position = "bottom") +
  scale_color_manual(values = c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF")) +
  labs( title = "", color = "" )

bray_visit_chg_plot
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 8) # Adjust dimensions as needed
bray_visit_chg_plot
dev.off()

#Save plot as pdf
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 2000, res = 300)
bray_visit_chg_plot
dev.off()
```


#Water

#Prepare data
```{r}
#Ps object with exact counts
ps_bds<-water_rare_ps

#reformat label
sample_data(ps_bds)$visit_2 <- fct_recode(sample_data(ps_bds)$vcode_prevfootcomp, 
                                        "Pre-trt" = "1", 
                                        "3 Months" = "3", 
                                        "12 Months" = "6",
                                        "Post-trt" = "7")

#set testing parameter
x="visit_2"

#Import tree
phy_tree(ps_bds)<-qza_to_phyloseq(tree="/Users/michellebode/maryland/16S/processed_data/tree/rooted_tree.qza")
ps_bds

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot

print("permanova")
permanova_result <- vegan::adonis2(bray_dist ~ visit_2, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)

bray_visit_water<-ps_bds
```

#Graph
```{r}
y<-"Bray_Curtis_visit_water"


# Run NMDS 
ordination_result <- ordinate(bray_visit_water, method='NMDS', distance='bray', k=4)

#Check stress levels if >0.2 increase k. ideally <0.1
ordination_result$stress

# Plot Axis 1 and Axis 2
bray_visit_water_plot<-plot_ordination(bray_visit_water, ordination_result, axes=c(1,2), color=x)+
geom_point(size = 6) +
  geom_point(
    size = 6,
    shape = 21,        
    color = "black",   
    stroke = 0.3) +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
))) +
  theme_classic()+
  theme_bw()+
     theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.title =element_text(size=26),
       legend.position = "bottom") +
  scale_color_manual(values = c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF")) +
  labs( title = "", color = "" )

bray_visit_water_plot
```

```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 8) # Adjust dimensions as needed
bray_visit_water_plot
dev.off()

#Save plot as pdf
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 2000, res = 300)
bray_visit_water_plot
dev.off()
```

#Combine Graphs
```{r}
x<-"combined_BC_visit"

# p1 has legend removed
p1 <- bray_visit_chg_plot + 
  theme(legend.position = "none") + 
  labs(title = "Chlorhexidine treatment") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.margin = margin(b = 30))

# p2 keeps legend
p2 <- bray_visit_water_plot + 
   theme(legend.position = "bottom",legend.margin = margin(t = 15), plot.margin = margin(t = 30)
  )+
  labs(title = "Soap and water treatment") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

combined_BC_visit <- p1 / p2 

combined_BC_visit
```

```{r}
#Save plot as pdf

setwd(f)
pdf(file = paste0(x,".pdf"), width = 10.333, height =  7.5) 
combined_BC_visit
dev.off()

#Save as image
png(filename = paste0(x,".png"), width =  3000 , height = 5400, res = 300)
combined_BC_visit
dev.off()
```


#Supplemental Figure 3C: Shift from visit 1 to visit 3

#isolate treatment groups
```{r}
v1v3<-subset_samples(ps_rare_genus, vcode_prevfootcomp %in% c("1","3"))
ps_rare_water<-subset_samples(v1v3, treatgroup == "0")
ps_rare_chg<-subset_samples(v1v3, treatgroup == "1")

# Keep only Patient_id that have both visit 1 and 3
meta_c<-as.data.frame(sample_data(ps_rare_chg))
valid_patients_chg <- meta_c %>%
  group_by(Patient_id) %>%
  filter(all(c("1", "3") %in% vcode_prevfootcomp)) %>%   # visit is probably character, so use "1","3"
  pull(Patient_id) %>%                      # <- extract as vector
  unique()

ps_rare_chg <- prune_samples(sample_names(ps_rare_chg)[sample_data(ps_rare_chg)$Patient_id %in% valid_patients_chg],ps_rare_chg)

meta_w<-as.data.frame(sample_data(ps_rare_water))
valid_patients_water <- meta_w %>%
  group_by(Patient_id) %>%
  filter(all(c("1", "3") %in% vcode_prevfootcomp)) %>%   # visit is probably character, so use "1","3"
  pull(Patient_id) %>%                      # <- extract as vector
  unique()

ps_rare_water <- prune_samples(sample_names(ps_rare_water)[sample_data(ps_rare_water)$Patient_id %in% valid_patients_water],ps_rare_water)

ps_rare_chg
ps_rare_water

#determine number of patients
length(unique(sample_data(ps_rare_chg)$Patient_id))
length(unique(sample_data(ps_rare_water)$Patient_id))
```

#CHG

#Prepare data
```{r}
#transform to relative abundance 
rel_abundance = transform_sample_counts(ps_rare_chg, function(x) x*100/ sum(x))

#Ps object with relative abundance
ps_bds<-rel_abundance 

#Ps object with exact counts
#ps_bds<-unique_genus

#set testing parameter
x="vcode_prevfootcomp"

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

print("permanova")
#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot
bray_dist <- distance(ps_bds, method = "bray")
permanova_result <- vegan::adonis2(bray_dist ~ vcode_prevfootcomp, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)
```

#Graphing differences
```{r}
y <- "Bray_Curtis_NMDS_Chg_1v3"

#order
meta <-as.data.frame(sample_data(ps_bds))
meta$visit <- factor(meta$visit, levels = c("1", "3"), ordered = TRUE)
sample_data(ps_bds) <- meta

# Run NMDS 
ordination_result <- ordinate(ps_bds, method = 'NMDS', distance = 'bray', k = 4)

# Check stress
ordination_result$stress

# Plot Axis 1 and Axis 2
Bray_Curtis_NMDS_Chg_1v3<- plot_ordination(ps_bds, ordination_result, axes = c(1,2), color = "visit_2") +
  geom_point(size =6, 
             alpha =1)+
  geom_point(
    size = 6,
    alpha = 1,
    shape = 21,
    stroke = 0.3,
    color ="black") +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#6BE0F2","#FFF86E"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
  )))+
  geom_path(
    aes(group = Patient_id),
    data = function(df) df[order(df$Patient_id, df$visit_2), ],
    arrow = arrow(type = "closed", length = unit(0.2, "inches")),
    linewidth = 0.4,
    color = "black",
    alpha = 0.9
  ) +
 theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
       legend.position = "none",
        legend.text =element_text(size=26,face ="italic"),
       legend.title = element_text(size=26))+
  labs(title = "", color = "")+
  scale_color_manual(values = c("Pre-treatment"="#6BE0F2", "3 Months"="#FFF86E"))


Bray_Curtis_NMDS_Chg_1v3
```

#print graphs
```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 6) # Adjust dimensions as needed
Bray_Curtis_NMDS_Chg_1v3 # Replace with your plot object
dev.off()

#Save plot as pdf
setwd(f)
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 1500, res = 300)
Bray_Curtis_NMDS_Chg_1v3  # Replace with your actual graph object
dev.off()
```


#Water

#Prepare data
```{r}
#transform to relative abundance 
rel_abundance = transform_sample_counts(ps_rare_water, function(x) x*100/ sum(x))

#Ps object with relative abundance
ps_bds<-rel_abundance 

#Ps object with exact counts
#ps_bds<-unique_genus

#set testing parameter
x="vcode_prevfootcomp"

#statistically test for the observed (or unobserved) difference between the groups using permanova. 
#But before running permanova, test and make sure that the assumption of homogenity of dispersion is met as this can affect your permanova results.
#If the assumption is not met (ie the dispersion test is significant), you can still do permanova but have in mind that if it is significant, the differences might be due to unequal dispersion between the groups and not their centroids (average microbial community composition)

meta <- as(sample_data(ps_bds), "data.frame")
meta_clean <- meta[!is.na(meta[[x]]), ] 
sample_data(ps_bds)<-meta_clean

#check dispersion assumption, if pval is non-significant it means that the assumption of homogenity of dispersion holds and we can proceed with permanova

bray_dist <- distance(ps_bds, method = "bray")
dispersion_result <- vegan::betadisper(bray_dist, sample_data(ps_bds)[[x]]) 

print("anova")
anova(dispersion_result)

print("TurkeyHSD")
TukeyHSD(dispersion_result)

#PERMDISP - confirms dispersion assumption. Basically performs permutataion tests on the dispesion results from betadisper, should be non-significant too to meet the assumption.

print("PERMDISP" )
vegan::permutest(dispersion_result, permutations = 999)

print("permanova")
#In permanova result, pay attention to the p-value and Rsquare value (effect size). If the effect size is very small, this could be the reason why the difference is not visible in the PCOA plot
bray_dist <- distance(ps_bds, method = "bray")
permanova_result <- vegan::adonis2(bray_dist ~ vcode_prevfootcomp, data = as(sample_data(ps_bds), "data.frame"))
print(permanova_result)
```

#Graphing differences
```{r}
y <- "Bray_Curtis_NMDS_water_1v3"

#order
meta <-as.data.frame(sample_data(ps_bds))
meta$visit <- factor(meta$visit, levels = c("1", "3"), ordered = TRUE)
sample_data(ps_bds) <- meta

# Run NMDS 
ordination_result <- ordinate(ps_bds, method = 'NMDS', distance = 'bray', k = 4)

# Check stress
ordination_result$stress

# Plot Axis 1 and Axis 2
Bray_Curtis_NMDS_water_1v3<- plot_ordination(ps_bds, ordination_result, axes = c(1,2), color = "visit_2") +
  geom_point(size =6, 
             alpha =1)+
  geom_point(
    size = 6,
    alpha = 1,
    shape = 21,
    stroke = 0.3,
    color ="black") +
  guides(color = guide_legend(
  override.aes = list(
    shape = 21,
    fill = c("#6BE0F2","#FFF86E"), # match your fill colors
    color = "black",  # dark border
    stroke = 0.4,
    size = 6
  )))+
  geom_path(
    aes(group = Patient_id),
    data = function(df) df[order(df$Patient_id, df$visit_2), ],
    arrow = arrow(type = "closed", length = unit(0.2, "inches")),
    linewidth = 0.4,
    color = "black",
    alpha = 0.9
  ) +
 theme_classic(base_size = 26)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
       legend.position = "none",
       legend.title = element_text(size=26),
        legend.text =element_text(size=26,face ="italic"))+
  labs(title = "", color = "")+
  scale_color_manual(values = c("Pre-treatment"="#6BE0F2","3 Months"="#FFF86E"))


Bray_Curtis_NMDS_water_1v3
```

#print graphs
```{r}
#Save plot as pdf
setwd(f)
pdf(file = paste0(y, "_axis_1_and_2.pdf"), width = 10, height = 6) 
Bray_Curtis_NMDS_water_1v3
dev.off()

#Save plot as pdf
setwd(f)
png(filename = paste0(y, "_axis_1_and_2.png"), width = 2500, height = 1500, res = 300)
Bray_Curtis_NMDS_water_1v3
dev.off()
```

#Combine Graphs
```{r}
x<-"combined_v1_v3"

# p1 has legend removed
p1 <- Bray_Curtis_NMDS_Chg_1v3 + 
  theme(legend.position = "none") + 
  labs(title = "Chlorhexidine treatment") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.margin = margin(b = 30))

# p2 keeps legend
p2 <- Bray_Curtis_NMDS_water_1v3 + 
   theme(legend.position = "bottom",legend.margin = margin(t = 15), plot.margin = margin(t = 30)
  )+
  labs(title = "Soap and water treatment") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

combined_v1_v3 <- p1 / p2 

combined_v1_v3

```

```{r}
#Save plot as pdf

setwd(f)
pdf(file = paste0(x,".pdf"), width = 10.333, height =  7.5) 
combined_v1_v3
dev.off()

#Save as image
png(filename = paste0(x,".png"), width =  3000 , height = 5400, res = 300)
combined_v1_v3
dev.off()
```


#Merge supplemental figure 3 plots
```{r}
x <-"supplemental_figure_3_merged"


bioburden_boxplot_chg_2 <-(bioburden_boxplot_chg/bioburden_boxplot_water) + theme(plot.margin = margin(b = 0,
                                                                              t = 50,
                                                                              r = 30,
                                                                              l =50))


combined_BC_visit_2 <- combined_BC_visit + theme(plot.margin = margin(b = 30,
                                            t = 30,
                                            l = 50,
                                            r = 50)) 

combined_v1_v3_2 <- combined_v1_v3 + theme(plot.margin = margin(b = 30,
                                            t = 30,
                                            l = 50,
                                            r = 50)) 

supplemental_figure_3<- (
  bioburden_boxplot_chg_2 | 
  combined_BC_visit_2 |
  combined_v1_v3_2 
) + 
  plot_layout(widths = c(1,1,1))


#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 31.333, height =  18) 
supplemental_figure_3
dev.off()
#Save as image
png(filename = paste0(x,".png"), width =  9400 , height = 5400, res = 300)
supplemental_figure_3
dev.off()
```


#Supplemental figure 4


#Supplemental figure 4A:Staph boxplot

#Graph of CHG bioburden box plot seperate 
```{r}
#Order factors correctly
df_staph_chg$visit_2 <- factor(df_staph_chg$visit_2, levels = c("Pre-treatment","3 Months","12 Months", "Post-treatment"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment","3 Months"),c("Pre-treatment","12 Months"),c("Pre-treatment","Post-treatment"))

#Check if adjusted p-values align with results determined later
stats<-compare_means(Abundance~ visit_2,
              data = df_staph_chg,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

y <-"Staph_chg_boxplot"

#Visualization 

library(ggplot2)

staph_boxplot_chg<-ggplot(df_staph_chg, aes(x=visit_2, y=Abundance, fill=visit_2))+
  geom_jitter(
  aes(fill = visit_2),
  shape = 21,
  color = "black",
  stroke = 0.3,
  width = 0.15,
  size = 2,
  alpha = 0.5)+
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)  +
  coord_cartesian(ylim = c(0, 135), clip = "off")+
  scale_x_discrete(labels =c(
  "Pre-treatment" = "Pre-trt\n(n=42)",  
  "3 Months" = "3 Months\n(n=27)",
  "12 Months" = "12 Months\n(n=25)",
  "Post-treatment" = "Post-trt\n(n=28)"))+
  theme_classic(base_size = 20)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.position = "none",
        legend.title =element_text(size=26))+
labs(title = "", x = "Chlorhexidine treatment", y = expression(bold("Relative abundance of Staphylococcus %")))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     label.y = c(100,117,134),
                     size = 12,
                     comparisons = comparisons)
  
staph_boxplot_chg
```

```{r}
setwd(f)

#Save as image
png(filename = paste0(y,".png"), width = 3000, height = 2000, res = 300)
staph_boxplot_chg # Replace with your actual graph object
dev.off()

#Save plot as pdf
pdf(file = paste0(y,".pdf"), width = 14, height = 7) # Adjust dimensions as needed
staph_boxplot_chg # Replace with your plot object
dev.off()
```


#Average Abundance
```{r}
interest_group<-df_staph_chg

summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean_conc = mean(Abundance, na.rm = TRUE),
    sd_conc = sd(Abundance, na.rm = TRUE),
    se = sd_conc / sqrt(n),
    ci95_low  = mean_conc - qt(0.975, df = n-1) * se,
    ci95_high = mean_conc + qt(0.975, df = n-1) * se
  )

summary_stats
```



#Graph of water bioburden box plot seperate 
```{r}
#Order factors correctly
df_staph_water$visit_2 <- factor(df_staph_water$visit_2, levels = c("Pre-treatment","3 Months","12 Months", "Post-treatment"))

#Perform Wilcox test on:
comparisons<-list(c("Pre-treatment","3 Months"),c("Pre-treatment","12 Months"),c("Pre-treatment","Post-treatment"))

#Check if adjusted p-values align with results determined later
stats<-compare_means(Abundance~ visit_2,
              data = df_staph_water,
              method = "wilcox",
              p.adjust.method = "holm",
              comparisons = comparisons)
stats

y <-"Staph_water_boxplot"

#Visualization 

library(ggplot2)

staph_boxplot_water<-ggplot(df_staph_water, aes(x=visit_2, y=Abundance, fill=visit_2))+
  geom_jitter(
  aes(fill = visit_2),
  shape = 21,
  color = "black",
  stroke = 0.3,
  width = 0.15,
  size = 2,
  alpha = 0.5)+
geom_boxplot(
  aes(fill = visit_2),
  color = "black",
  size = 0.3,       
  alpha = 0.5,
  outlier.shape = NA,
)  +
  coord_cartesian(ylim = c(0, 135), clip = "off")+
  scale_x_discrete(labels =c(
  "Pre-treatment" = "Pre-trt\n(n=27)",  
  "3 Months" = "3 Months\n(n=24)",
  "12 Months" = "12 Months\n(n=22)",
  "Post-treatment" = "Post-trt\n(n=24)"))+
  theme_classic(base_size = 20)+

 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.position = "none",
        legend.title =element_text(size=26))+
labs(title = "", x = "Soap and water treatment", y = expression(bold("Relative abundance of Staphylococcus %")))+
  scale_fill_manual(values=c("#6BE0F2","#FFF86E","#FCC44E","#69F5AF"))+
  stat_compare_means(method = "wilcox",
                     label = "p.signif",
                     label.y = c(100,117,134),
                     size = 12,
                     comparisons = comparisons)
  
staph_boxplot_water
```

```{r}
setwd(f)

#Save as image
png(filename = paste0(y,".png"), width = 3000, height = 2000, res = 300)
staph_boxplot_water # Replace with your actual graph object
dev.off()

#Save plot as pdf
pdf(file = paste0(y,".pdf"), width = 14, height = 7) # Adjust dimensions as needed
staph_boxplot_water # Replace with your plot object
dev.off()
```

#Average Abundance
```{r}
interest_group<-df_staph_water

summary_stats <- interest_group %>%
  group_by(visit_2) %>%
  summarise(
    n = n(),
    mean_conc = mean(Abundance, na.rm = TRUE),
    sd_conc = sd(Abundance, na.rm = TRUE),
    se = sd_conc / sqrt(n),
    ci95_low  = mean_conc - qt(0.975, df = n-1) * se,
    ci95_high = mean_conc + qt(0.975, df = n-1) * se
  )

summary_stats
```

#Supplemental figure 4B:Staph compared to bioburden

#bioburden and staph corrleation
```{r}
correlation_water<-df_staph_water
correlation_chg<-df_staph_chg

merge_water_group<-merge(grouped_water_dpcr, correlation_water, by.x="SAMPLE_NAME", by.y="Sample")
merge_chg_group<-merge(grouped_chg_dpcr, correlation_chg, by.x="SAMPLE_NAME", by.y="Sample")
```

```{r}
staph_bioburden_chg<-ggplot(merge_chg_group, aes(y=log_Concentration, x=Abundance))+
  #geom_line(colour = "black")+
  geom_point(size = 3,
             colour = "#557689")+
    geom_smooth(
    data = merge_chg_group,
    aes(y = log_Concentration, x = Abundance),
    method = "lm",
    se = TRUE,
    color = "#557689"
  ) +

  ## correlation
  stat_cor(method = "spearman",
    data = merge_chg_group,
    aes(y = log_Concentration, x = Abundance),
    label.x.npc = "middle",
    label.y.npc = "right",
    size = 6
  ) +
  geom_point(colour = "black", shape = 21, size = 3)+
  #scale_colour_manual(values=c("Pre-treatment"="#6BE0F2","3 Months"="#FFF86E","12 Months"="#FCC44E","Post-treatment"="#69F5AF","Increase" = "#05A106", "Decrease" ="#D10A0A", "No change"="#A8A7A7"), breaks =c("Increase","Decrease"), guide = guide_legend(override.aes =list(linewidth =2, size = 12)))+
  labs(title = "",  y = expression(bold("log"[10]* " Concentration")),, x = expression(bold("Relative abundance of Staphylococcus %")))+
  theme_classic(base_size = 20)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.position = "bottom",
        legend.title =element_text(size=26))

staph_bioburden_chg
```

```{r}
staph_bioburden_water<-ggplot(merge_water_group, aes(y=log_Concentration, x=Abundance))+
  #geom_line(colour = "black")+
  geom_point(size = 3,
             colour = "#557689")+
    geom_smooth(
    data = merge_water_group,
    aes(y = log_Concentration, x = Abundance),
    method = "lm",
    se = TRUE,
    color = "#557689"
  ) +

  ## correlation
  stat_cor(method = "spearman",
    data = merge_water_group,
    aes(y = log_Concentration, x = Abundance),
    label.x.npc = "right",
    label.y.npc = "bottom",
    size = 6
  ) +
  geom_point(colour = "black", shape = 21, size = 3)+
  #scale_colour_manual(values=c("Pre-treatment"="#6BE0F2","3 Months"="#FFF86E","12 Months"="#FCC44E","Post-treatment"="#69F5AF","Increase" = "#05A106", "Decrease" ="#D10A0A", "No change"="#A8A7A7"), breaks =c("Increase","Decrease"), guide = guide_legend(override.aes =list(linewidth =2, size = 12)))+
  labs(title = "",  y = expression(bold("log"[10]* " Concentration")),, x = expression(bold("Relative abundance of Staphylococcus %")))+
  theme_classic(base_size = 20)+
 theme(strip.background = element_blank(), 
        axis.text.y = element_text(size=20),
        axis.title.y = element_text(size=26, vjust=2, face = "bold"),
        axis.title.x.bottom = element_text(size=26,vjust=-1, face = "bold"),
        axis.text.x.bottom = element_text(angle = 0,size = 20),
        legend.text =element_text(size=26,face ="italic"),
        legend.position = "bottom",
        legend.title =element_text(size=26))

staph_bioburden_water
```

#Combine Graphs
```{r}
x<-"combined_staph_bioburden"

# p1 has legend removed
p1 <- staph_bioburden_chg + 
  theme(legend.position = "none") + 
  labs(title = "Chlorhexidine treatment") + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.margin = margin(b = 30))

# p2 keeps legend
p2 <- staph_bioburden_water + 
   theme(legend.position = "bottom",legend.margin = margin(t = 15), plot.margin = margin(t = 30)
  )+
  labs(title = "Soap and water treatment") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

combined_staph_bioburden <- p1 / p2 

combined_staph_bioburden
```

```{r}
#Save plot as pdf

setwd(f)
pdf(file = paste0(x,".pdf"), width = 10.333, height =  7.5) 
combined_staph_bioburden
dev.off()

#Save as image
png(filename = paste0(x,".png"), width =  3000 , height = 5400, res = 300)
combined_staph_bioburden
dev.off()
```

#Merge supplemental figure 4 plots
```{r}
x <-"supplemental_figure_4_merged"


staph_boxplot_2 <-(staph_boxplot_chg/staph_boxplot_water) + theme(plot.margin = margin(b = 0,
                                                                              t = 50,
                                                                              r = 30,
                                                                              l =50))

combined_staph_bioburden_2 <-combined_staph_bioburden + theme(plot.margin = margin(b = 0,
                                                                              t = 50,
                                                                              r = 30,
                                                                              l =50))

supplemental_figure_4<- (
  staph_boxplot_2 | 
  combined_staph_bioburden_2
) + 
  plot_layout(widths = c(1,1))


#Save plot as pdf
setwd(f)
pdf(file = paste0(x,".pdf"), width = 31.333, height =  18) 
supplemental_figure_4
dev.off()
#Save as image
png(filename = paste0(x,".png"), width =  9400 , height = 5400, res = 300)
supplemental_figure_4
dev.off()
```




