---
title: "Maryland_ITS_Analysis_Prep_Final"
author: "Michelle Bode"
output: html_document
date: "2026-01-06"
---

#Set working directory
```{r setup, include=TRUE}
knitr::opts_knit$set(root.dir = "[input pathway to directory with data]")

#Directory for saving figures
f<-"[input pathway to where figure will be saved]"
```

#Install required packages
```{r}
#BiocManager::install("phyloseq")
#install.packages("dplyr")
#install.packages("ggpubr")
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("BiocManager")
#BiocManager::install("ALDEx2")
#BiocManager::install("decontam")
#install.packages(c("devtools", "httr", "jsonlite", "tidyverse", "reticulate"))
#devtools::install_github("jbisanz/qiime2R")
#install.packages("vegan")
#BiocManager::install("decontam")
```

#Activate Packages
```{r}
library("ggplot2")
library("patchwork")
library("ggpubr")
library("ALDEx2")
library("phyloseq")
library("dplyr")
library("tidyverse")
library ("ggplot2")
library ("BiocManager")
library("devtools")
library("httr")
library("jsonlite") 
library("tidyverse")
library("reticulate")
library("qiime2R")
library("vegan")
library("decontam")
```

#Import data
```{r}
#Import raw data
tax_table<-as.matrix(read.csv2("taxonomy_table_ITS.csv", sep=",", row.names=1))
otu_table<-read.csv2("otu_table_ITS.csv", sep=",", row.names=1, check.names = FALSE)
metadata_table<-read.csv2("metadata.csv", sep=",", row.names=1)

#Merge to a phyloseq object
ps_raw_ITS <- phyloseq(
  otu_table(otu_table,taxa_are_rows = TRUE),
  tax_table(tax_table),
  sample_data(metadata_table))
  
ps_raw_ITS 
```

#Remove samples that are not included in analysis
```{r}
#Remove patients with current complication on the same foot
ps_raw_analysis <-subset_samples(ps_raw_ITS, !Patient_id %in% c("1016","1037","1072","1194"))
ps_raw_analysis <-prune_samples(sample_sums(ps_raw_analysis)>0, ps_raw_analysis)

#view
ps_raw_analysis
```

#Summary stats of patient samples that will be used in analysis (before filtering)
```{r}
meta_raw <- sample_data(ps_raw_analysis)

#Total treatment group summary
print("chg")
chg<-subset(meta_raw, treatgroup == 1)
print("chg total samples")
nrow(chg)
print("chg total patients")
length(unique(chg$Patient_id))

print("water")
water<-subset(meta_raw, treatgroup == 0)
print("water total samples")
nrow(water)
print("water total patients")
length(unique(water$Patient_id))

#Summary at each visit 
print("chg visit breakdown")
meta_chg<-chg
print("pre-treatment")
nrow(subset(meta_chg, vcode_prevfootcomp == 1))
print("3 months")
nrow(subset(meta_chg, vcode_prevfootcomp == 3))
print("12 months")
nrow(subset(meta_chg, vcode_prevfootcomp == 6))
print("post-treatment")
nrow(subset(meta_chg, vcode_prevfootcomp == 7))

print("water visit breakdown")
meta_water<-water
print("pre-treatment")
nrow(subset(meta_water, vcode_prevfootcomp == 1))
print("3 months")
nrow(subset(meta_water, vcode_prevfootcomp == 3))
print("12 months")
nrow(subset(meta_water, vcode_prevfootcomp == 6))
print("post-treatment")
nrow(subset(meta_water, vcode_prevfootcomp == 7))
```

#format metadata
```{r}
# Extract and convert sample_data to data frame
sam_data <- as.data.frame(sample_data(ps_raw_analysis))

#change names to right string vector. For example 1 vs 0 should be read as characters and not integers because the values aren't continuous they represent variables.
sam_data[,c(1:5,7:12,14:35)] <-lapply(sam_data[,c(1:5,7:12,14:35)], as.character)

#Change number representations into label names for figures
sam_data$treatgroup_2<-ifelse(sam_data[["treatgroup"]] %in% c("0"),"Soap and water","Chlorhexidine")

sam_data$treatment_active <- ifelse(sam_data$vcode_prevfootcomp %in% c("1","7"),"No active treatment (pre-trt + post-trt)","Active treatment (3 months + 12 months)")

#Rename visits
sam_data$visit_2 <- fct_recode(sam_data$vcode_prevfootcomp, 
                                        "Pre-treatment" = "1", 
                                        "3 Months" = "3", 
                                        "12 Months" = "6",
                                        "Post-treatment" = "7")

#Update phyloseq object
ps_clean_metadata<-ps_raw_analysis
sample_data(ps_clean_metadata)<-as.data.frame(sam_data)
```

#View if top distributed taxa are identified
```{r}
#Isolate otu tables
feature_counts<-data.frame(otu_table(ps_clean_metadata))

#Count cells in rows greater than zero (ie. how many samples have that feature)
clean_row<-rowSums(feature_counts > 0)

#Organize from greatest to least
ordered_feautres <- clean_row[order(-clean_row)]

#Check data
ordered_feautres

#Convert to data.frame
counts_df <-data.frame(sample_counts = ordered_feautres)

#View
counts_df

#isolate taxonomy data
taxonomy_data<-data.frame(tax_table(ps_clean_metadata))

#merge taxonomy and otu data
all_wound_count<-merge(taxonomy_data,counts_df, by="row.names", all=TRUE)

#make column headers a row
colnames(all_wound_count)[1]<-"OTU ID"

#remove na
clean_wound_count <- all_wound_count %>% filter(!is.na(sample_counts))

#Order by sample count
ordered_clean_wound<-clean_wound_count[order(-clean_wound_count$sample_counts),]

#Check samples
ordered_clean_wound
```

#Add genus names for missing top taxa (identified using BLAST+ on the NCBI website)
```{r}
# Create the OTU-to-Genus mapping data frame
otu_to_genus <- data.frame(
  OTU_ID = c("21b7198768350673f6373f807a0dc3f6","403fdd28fac726454c161ffb1d3a393d","81aa1b36cd9303696bc03435d0947bb1","9150863e1ac79fa37b4a2f74a4326f18","d0c33b0f89767b49ea3ac2673609c39e","26de6680481d022ad25fda417b89e40a","74f9ddbe696b88d5ea6a327519622423","6c1c9d1e92b43cf4e54c083ec3a15f37"),
  Genus = c("Candidozyma"," Chaetomium","f_Ustilaginaceae","o_Saccharomycetales","Malassezia", "c_Sordariomycetes","o_Chaetothyriales","o_Agaricales"))

# Extract the taxonomy table from the phyloseq object
tax_table_df <- as.data.frame(tax_table(ps_clean_metadata))

# Update the Genus column based on the OTU-to-Genus mapping
tax_table_df$Genus <- ifelse(
  rownames(tax_table_df) %in% otu_to_genus$OTU_ID, 
  otu_to_genus$Genus[match(rownames(tax_table_df), otu_to_genus$OTU_ID)], 
  tax_table_df$Genus
)

# Replace the updated taxonomy table in the phyloseq object
tax_table(ps_clean_metadata) <- as.matrix(tax_table_df)

# View the updated taxonomy table for the specific OTUs
tax_table_df[otu_to_genus$OTU_ID, ]

ps_clean_metadata
```

#Only keep fungi 
```{r}
#create copy with new name
ps_fungi<-ps_clean_metadata

# Access the sample data
taxonomy <- as.data.frame(tax_table(ps_fungi))

#See how many are not fungi 
taxonomy_non_fungi <-subset(taxonomy, Kingdom != "Fungi")
print("How many non fungi ")
nrow(taxonomy_non_fungi )
print("summary of reads lost")
ps_non_fungi<-subset_taxa(ps_fungi, Kingdom != "Fungi")
summary(sample_sums(ps_non_fungi))
summary(taxa_sums(ps_non_fungi))

# Keep only samples identified as fungi 
taxonomy_filtered <-subset(taxonomy, Kingdom == "Fungi")

#see how many were lost
print("How many removed")
nrow(taxonomy) - nrow(taxonomy_filtered)

#update phyloseq
tax_table(ps_fungi )<-as.matrix(taxonomy_filtered)

#view
ps_fungi 
```

#View distribution of taxonomic identification
```{r}
library(tibble)
unidentifed_taxa<-tax_table(ps_fungi )

#counts total number identified at each taxonomic level
identifed <- colSums(!is.na(unidentifed_taxa))

#labels counts as Identified and makes Taxonomic level a column
labeled_counts<-data.frame(Identified = identifed)
taxonomy_set <-rownames_to_column(labeled_counts, var="Taxonomic Level")
taxonomy_set

# Extract the value for the "Kingdom" row in the "Identified" column
sum_value <- taxonomy_set$Identified[taxonomy_set$`Taxonomic Level`=="Kingdom"]

# calculate proportion identified
x <- taxonomy_set$Identified
taxonomy_set$`Percent Identified` <- x/sum_value*100 

#Ensures results are displayed in this order  
taxonomy_set$`Taxonomic Level` <- factor(taxonomy_set$`Taxonomic Level`, levels = c("Kingdom","Phylum","Class","Order","Family","Genus","Species"))

#Graph Results
taxonomy_graph <-ggplot(taxonomy_set, aes(x=`Taxonomic Level`, y= `Percent Identified`))+
  geom_bar(stat = "identity", fill="steelblue")+
  theme_classic()+
  labs(title = "Percent of features identified to each taxanomic level", x = "Taxonomic Level", y = "Percent Identified")+
  theme(strip.background = element_blank(), axis.text.x.bottom = element_text(angle = -90))

taxonomy_graph 


#Save graph

setwd(f)

#Save pdf
pdf(file = "Percent_of_features_identified_to_each_taxanomic_level.pdf",width = 10, height = 6) 
print(taxonomy_graph) # Replace with your plot object
dev.off()

#Save png
png(file = "Percent_of_features_identified_to_each_taxanomic_level.png", width = 2500, height = 1500, res = 300)

print(taxonomy_graph) # Replace with your plot object
dev.off()
```

#Remove empty samples and low abundant
```{r}
#Remove empty samples and low abundant taxa
remove_empty_1_ps<-ps_fungi 

#Remove empty samples
remove_empty_1_ps<-prune_samples(sample_sums(remove_empty_1_ps) > 0, remove_empty_1_ps)


#Remove very low abundant features 
remove_low_ps<-prune_taxa(taxa_sums(remove_empty_1_ps)>5, remove_empty_1_ps) 
remove_low_ps
```

#Filter out contaminants 
```{r}
ps_filter<-remove_low_ps

#Separate metadata
sample_data<-sample_data(ps_filter)

#Create column with row_names
sample_data$Sample_name<-rownames(sample_data)

#Replace negative controls with samples with NA
sample_data$Sample_name<-gsub("NEG",NA,sample_data$Sample_name)
sample_data$Sample_name<-gsub("NC",NA,sample_data$Sample_name)
sample_data$Sample_name[is.na(sample_data$Sample_name)] <- "Neg"

#Update phyloseq
sample_data(ps_filter)<-sample_data

#Identifies all negative controls and created a "is.neg" column
sample_data(ps_filter)$is.neg <- sample_data(ps_filter)$Sample_name == "Neg"

#filters contaminants based on prevalence in negative control compared to samples
contamdf.prev <- isContaminant(ps_filter, method="prevalence", neg= "is.neg")

print("Number of contaminant taxa")
table(contamdf.prev$contaminant)

#prune taxa to only include noncontaminated sequences
print("After contaminant cleaning")
ps.noncontam <- prune_taxa(!contamdf.prev$contaminant, ps_filter)
ps.noncontam #? taxa
```

#View Contaminanted Taxa
```{r}
# Extract the contaminant taxa (those labeled as TRUE)
contaminant_taxa <- rownames(contamdf.prev[contamdf.prev$contaminant == TRUE, ])

# Extract taxonomy table for the contaminant taxa
tax_table_contaminants <- tax_table(ps_filter)[contaminant_taxa, ]

#view counts of contaminants before removal
clean_otu<-otu_table(remove_low_ps)
names<-rownames(tax_table_contaminants)
count_present<-clean_otu[rownames(clean_otu) %in% names, ]
total_count_present<-rowSums(count_present)
summary(total_count_present)

# View the taxonomy of the contaminants
print("contaminany names")
tax_table_contaminants
total_count_present
```

#Remove negative and positive controls from analysis
```{r}
# Create a new phyloseq object without negative controls
ps_clean <- prune_samples(!is.na(sample_data(ps.noncontam)$Patient_id), ps.noncontam)

# Check the resulting phyloseq object
removed_neg_controls<-sample_data(ps_clean)$Sample_name

ps_clean 
```

#Remove empty samples 
```{r}
#Remove empty samples and low abundant taxa
remove_empty_2_ps<-ps_clean

#Remove empty samples
remove_empty_2_ps<-prune_samples(sample_sums(remove_empty_2_ps) > 0, remove_empty_2_ps)

remove_empty_2_ps
```


#Summary stats of patient samples that will be used in analysis (after filtering)
```{r}
#Subset analysis group
ps_clean_analysis <-subset_samples(remove_empty_2_ps, prevfootcompanalysis == "1")

ps_clean_analysis <-prune_samples(sample_sums(ps_clean_analysis)>0, ps_clean_analysis)
meta_raw <- sample_data(ps_clean_analysis)

#Total treatment group summary
print("chg")
chg<-subset(meta_raw, treatgroup == 1)
print("chg total samples")
nrow(chg)
print("chg total patients")
length(unique(chg$Patient_id))

print("water")
water<-subset(meta_raw, treatgroup == 0)
print("water total samples")
nrow(water)
print("water total patients")
length(unique(water$Patient_id))

#Summary at each visit 
print("chg visit breakdown")
meta_chg<-chg
print("pre-treatment")
nrow(subset(meta_chg, vcode_prevfootcomp == 1))
print("3 months")
nrow(subset(meta_chg, vcode_prevfootcomp == 3))
print("12 months")
nrow(subset(meta_chg, vcode_prevfootcomp == 6))
print("post-treatment")
nrow(subset(meta_chg, vcode_prevfootcomp == 7))

print("water visit breakdown")
meta_water<-water
print("pre-treatment")
nrow(subset(meta_water, vcode_prevfootcomp == 1))
print("3 months")
nrow(subset(meta_water, vcode_prevfootcomp == 3))
print("12 months")
nrow(subset(meta_water, vcode_prevfootcomp == 6))
print("post-treatment")
nrow(subset(meta_water, vcode_prevfootcomp == 7))

#Summarize sample size
summary(sample_sums(ps_clean_analysis))

```

#Prepare for normalization
```{r}
#Pick data type for analysis 
ps_all_samples <-ps_clean_analysis

#Isolate OTU Table 
otus <-as.matrix(otu_table(ps_all_samples))

#switch column and header
flipped_otu <- t(otus)
flipped_clean_otus <-as.data.frame(flipped_otu)
flipped_clean_otus

#generate rarefaction curves
vegan::rarecurve(flipped_clean_otus, step = 5, col = "blue", cex = 0.5, label =FALSE, ylim =c(0,40), xlim =c(0,4000))
```

#Removal for rarefraction minimum
```{r}

#remove samples below 
clean_otu <- otus[,colSums(otus) > 700 ]

#look at count distribution in samples after removal
clean_otu_stats<-summary(colSums(clean_otu))

print("distribution of features in samples")
clean_otu_stats

print("percent of samples maintained")
ncol(clean_otu)
ncol(clean_otu)/ncol(otus)
```

#Check sample distribution (after rarefaction minimum cutoff)
```{r}
#Select sample for viewing
ps_min_removed<-ps_clean_analysis
otu_table(ps_min_removed)<-clean_otu
meta_raw <- sample_data(ps_min_removed)

#Total treatment group summary
print("chg")
chg<-subset(meta_raw, treatgroup == 1)
print("chg total samples")
nrow(chg)
print("chg total patients")
length(unique(chg$Patient_id))

print("water")
water<-subset(meta_raw, treatgroup == 0)
print("water total samples")
nrow(water)
print("water total patients")
length(unique(water$Patient_id))

#Summary at each visit 
print("chg visit breakdown")
meta_chg<-chg
print("pre-treatment")
nrow(subset(meta_chg, vcode_prevfootcomp == 1))
print("3 months")
nrow(subset(meta_chg, vcode_prevfootcomp == 3))
print("12 months")
nrow(subset(meta_chg, vcode_prevfootcomp == 6))
print("post-treatment")
nrow(subset(meta_chg, vcode_prevfootcomp == 7))

print("water visit breakdown")
meta_water<-water
print("pre-treatment")
nrow(subset(meta_water, vcode_prevfootcomp == 1))
print("3 months")
nrow(subset(meta_water, vcode_prevfootcomp == 3))
print("12 months")
nrow(subset(meta_water, vcode_prevfootcomp == 6))
print("post-treatment")
nrow(subset(meta_water, vcode_prevfootcomp == 7))

#Summarize sample size
summary(sample_sums(ps_min_removed))
```

#Remove empty features
```{r}
#Check number of empty features
sum(rowSums(clean_otu)<1)

#remove features (rows) with few counts
clean_otu_c <- clean_otu[rowSums(clean_otu)>0,]

#check with zeros removed
sum(rowSums(clean_otu_c)<1)

#Now check summary of features 
summary(rowSums(clean_otu_c))
```
#Save non-rarefied without minimum 
```{r}
ps_nr_min700<-ps_all_samples

otu_table(ps_nr_min700)<-clean_otu_c

print("ps_nr_min700")
ps_nr_min700
```

#Rarefraction
```{r}
#switch columns and header names
df_matrix <- as.matrix(clean_otu_c)
c_flipped_clean_otu <- t(df_matrix)
c_flipped_clean_otu<-as.data.frame(c_flipped_clean_otu)

#check changes
c_flipped_clean_otu

# Determine the minimum total number of individuals in any sample
min_sample_size <- min(rowSums(c_flipped_clean_otu))
min_sample_size

#Generates rarefied otu tables. With multiple different results.
rarefied_otu_data_1 <- vegan::rrarefy(c_flipped_clean_otu, min_sample_size)
rarefied_otu_data_2 <- vegan::rrarefy(c_flipped_clean_otu, min_sample_size)
rarefied_otu_data_3 <- vegan::rrarefy(c_flipped_clean_otu, min_sample_size)
```

#Remove empty features
```{r}
#Check empty feature count
sum(colSums(rarefied_otu_data_1 )<1)

#Remove empty features
rarefied_otu_data_1 <- rarefied_otu_data_1[,colSums(rarefied_otu_data_1)>0]

#Check empty feature count
sum(colSums(rarefied_otu_data_2)<1)

#Remove empty features
rarefied_otu_data_2 <- rarefied_otu_data_2 [,colSums(rarefied_otu_data_2)>0]

#Check empty feature count
sum(colSums(rarefied_otu_data_3)<1)

#Remove empty features
rarefied_otu_data_3 <- rarefied_otu_data_3 [,colSums(rarefied_otu_data_3)>0]
```

```{r}
#View summary of features
stats_rare_1<- colSums(rarefied_otu_data_1)
stats_rare_2<- colSums(rarefied_otu_data_2)
stats_rare_3<- colSums(rarefied_otu_data_3)

#View summary of samples
stats_raref_1<- rowSums(rarefied_otu_data_1)
stats_raref_2<- rowSums(rarefied_otu_data_2)
stats_raref_3<- rowSums(rarefied_otu_data_3)

print("feature summary")
summary(stats_rare_1)
summary(stats_rare_2)
summary(stats_rare_3)

print("sample summary")
summary(stats_raref_1)
summary(stats_raref_2)
summary(stats_raref_3)

```

#Updated phyloseq object
```{r}
#transfer rarefied data back to ps objects
ps_rare_1_ITS<-ps_all_samples
ps_rare_2_ITS<-ps_all_samples
ps_rare_3_ITS<-ps_all_samples

otu_1 <- t(rarefied_otu_data_1)
otu_1 <- otu_table(otu_1, taxa_are_rows = TRUE)
otu_1 -> otu_table(ps_rare_1_ITS) 

otu_2 <- t(rarefied_otu_data_2)
otu_2 <- otu_table(otu_2, taxa_are_rows = TRUE)
otu_2 -> otu_table(ps_rare_2_ITS) 

otu_3 <- t(rarefied_otu_data_3)
otu_3 <- otu_table(otu_3, taxa_are_rows = TRUE)
otu_3 -> otu_table(ps_rare_3_ITS) 

print("rarefraction_1")
ps_rare_1_ITS
print("rarefraction_2")
ps_rare_2_ITS
print("rarefraction_3")
ps_rare_2_ITS
```

#Merge to genus level taxa 

#Non-rarefaction data 
```{r}
#Add a different number for all NA in Genus
Genus_unknown<-ps_nr_min700

# Extract the taxonomy table from the phyloseq object
tax_table_cleaned <- as.data.frame(tax_table(Genus_unknown))

# Add a unique number to the NA values in the Phylum column
tax_table_cleaned$Genus[is.na(tax_table_cleaned$Genus)] <- paste("Unknown_genus", seq_along(tax_table_cleaned$Genus[is.na(tax_table_cleaned$Genus)]), sep="_")

# Update the taxonomy table in the phyloseq object
tax_table(Genus_unknown) <- as.matrix(tax_table_cleaned)

#Filter only unique Genus names (merge species with same names)
unique_genus<-tax_glom(Genus_unknown, taxrank="Genus")

ps_nr_genus_ITS<-unique_genus
ps_nr_genus_ITS
``` 


#Average OTU counts all non rarefied
```{r}
interest_group <- taxa_sums(ps_nr_min700)

summary_stats <- data.frame(value = interest_group) %>%
  summarise(n = n(),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    se = sd / sqrt(n),
    ci95_low  = mean - qt(0.975, df = n - 1) * se,
    ci95_high = mean + qt(0.975, df = n - 1) * se)

ps_nr_min700
summary_stats
```

#Average OTU counts all rarefied
```{r}
interest_group <- taxa_sums(ps_nr_genus_ITS)

summary_stats <- data.frame(value = interest_group) %>%
  summarise(n = n(),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    se = sd / sqrt(n),
    ci95_low  = mean - qt(0.975, df = n - 1) * se,
    ci95_high = mean + qt(0.975, df = n - 1) * se)

summary(interest_group)
ps_nr_genus_ITS
summary_stats
```

#Rarefaction data 
```{r}
#Add a different number for all NA in Genus
Genus_unknown<-ps_rare_1_ITS

# Extract the taxonomy table from the phyloseq object
tax_table_cleaned <- as.data.frame(tax_table(Genus_unknown))

# Add a unique number to the NA values in the Phylum column
tax_table_cleaned$Genus[is.na(tax_table_cleaned$Genus)] <- paste("Unknown_genus", seq_along(tax_table_cleaned$Genus[is.na(tax_table_cleaned$Genus)]), sep="_")

# Update the taxonomy table in the phyloseq object
tax_table(Genus_unknown) <- as.matrix(tax_table_cleaned)

#Filter only unique Genus names (merge species with same names)
unique_genus<-tax_glom(Genus_unknown, taxrank="Genus")

ps_rare_genus_ITS<-unique_genus
ps_rare_genus_ITS
``` 

